---
title: "rodent reproductive phenology (male)"
author: "Pat Dumandan"
date: "12/23/2020"
output: html_document
---

```{r warning=FALSE, echo=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

Portal_rodent=read.csv("./PortalData/Rodents/Portal_rodent.csv")
Portal_plots=read.csv("./PortalData/SiteandMethods/Portal_plots.csv")
portal1=inner_join(Portal_rodent, Portal_plots, by= c("plot", "month", "year"))%>%
  mutate(Month=recode(month, "1"= "Jan", "2"="Feb", "3"="Mar", "4"="Apr",
                      "5"="May","6"="Jun", "7"="Jul", "8"= "Aug", "9"="Sept",
                      "10"="Oct","11"="Nov", "12"="Dec"))
target_repro=c("S", "M", "R")
#male individuals####
#size thresholds####
PB=portal1%>%
  filter(species=="PB",testes %in% target_repro,wgt>=16)%>%
  arrange(wgt)

PP=portal1%>%
  filter(species=="PP",testes %in% target_repro, wgt>=10)%>%
  arrange(wgt)

PE=portal1%>%
  filter(species=="PE",testes %in% target_repro, wgt >=7)%>%
  arrange(wgt)

PF=portal1%>%
  filter(species=="PF",testes %in% target_repro, wgt>=4)%>%
  arrange(wgt)

PM=portal1%>%
  filter(species=="PM",testes %in% target_repro, wgt >=11)%>%
  arrange(wgt)
#1 individual identified as juv but size within reproductive??

RM=portal1%>%
  filter(species=="RM",testes %in% target_repro, wgt>=4)%>%
  arrange(wgt)

#### PB proportions#####

PBex=portal1%>%
  filter(species=="PB",testes %in% target_repro,wgt>=16, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

PBcon=portal1%>%
  filter(species=="PB",testes %in% target_repro,wgt>=16, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

PB_all_ex=portal1%>%
  filter(species=="PB", treatment=="exclosure", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

PB_all_con=portal1%>%
  filter(species=="PB", treatment=="control", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

pb1=left_join(PB_all_ex, PBex)
pb2=left_join(PB_all_con,PBcon)
pb3=rbind(pb1,pb2)

pb4=pb3%>%
  mutate(proportion=repro/count, species="PB")
write.csv(pb4, "PB_props.csv")
PB_props=read.csv("PB_props.csv")


#### PP proportions#####

PPex=portal1%>%
  filter(species=="PP",testes %in% target_repro,wgt>=10, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

PPcon=portal1%>%
  filter(species=="PP",testes %in% target_repro,wgt>=10, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

PP_all_ex=portal1%>%
  filter(species=="PP", treatment=="exclosure", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

PP_all_con=portal1%>%
  filter(species=="PP", treatment=="control", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

pp1=left_join(PP_all_ex, PPex)
pp2=left_join(PP_all_con,PPcon)
pp3=rbind(pp1,pp2)

pp4=pp3%>%
  mutate(proportion=repro/count, species="PP")%>%
  filter(proportion<=1)
write.csv(pp4, "PP_props.csv")
PP_props=read.csv("PP_props.csv")

#### PE proportions#####

PEex=portal1%>%
  filter(species=="PE",testes %in% target_repro,wgt>=7, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

PEcon=portal1%>%
  filter(species=="PE",testes %in% target_repro,wgt>=7, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

PE_all_ex=portal1%>%
  filter(species=="PE", treatment=="exclosure", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

PE_all_con=portal1%>%
  filter(species=="PE", treatment=="control", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

pe1=left_join(PE_all_ex, PEex)
pe2=left_join(PE_all_con,PEcon)
pe3=rbind(pe1,pe2)

pe4=pe3%>%
  mutate(proportion=repro/count, species="PE")
write.csv(pe4, "PE_props.csv")
PE_props=read.csv("PE_props.csv")

#### PF proportions#####

PFex=portal1%>%
  filter(species=="PF",testes %in% target_repro,wgt>=4, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

PFcon=portal1%>%
  filter(species=="PF",testes %in% target_repro,wgt>=4, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

PF_all_ex=portal1%>%
  filter(species=="PF", treatment=="exclosure", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

PF_all_con=portal1%>%
  filter(species=="PF", treatment=="control", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

pf1=left_join(PF_all_ex, PFex)
pf2=left_join(PF_all_con,PFcon)
pf3=rbind(pf1,pf2)

pf4=pf3%>%
  mutate(proportion=repro/count, species="PF")
write.csv(pf4, "PF_props.csv")
PF_props=read.csv("PF_props.csv")

#### PM proportions#####

PMex=portal1%>%
  filter(species=="PM",testes %in% target_repro,wgt>=11, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

PMcon=portal1%>%
  filter(species=="PM",testes %in% target_repro,wgt>=11, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

PM_all_ex=portal1%>%
  filter(species=="PM", treatment=="exclosure", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

PM_all_con=portal1%>%
  filter(species=="PM", treatment=="control", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

pm1=left_join(PM_all_ex, PMex)
pm2=left_join(PM_all_con,PMcon)
pm3=rbind(pm1,pm2)

pm4=pm3%>%
  mutate(proportion=repro/count, species="PM")
write.csv(pm4, "PM_props.csv")
PM_props=read.csv("PM_props.csv")

#### RM proportions#####

RMex=portal1%>%
  filter(species=="RM",testes %in% target_repro,wgt>=4, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

RMcon=portal1%>%
  filter(species=="RM",testes %in% target_repro,wgt>=4, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

RM_all_ex=portal1%>%
  filter(species=="RM", treatment=="exclosure", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

RM_all_con=portal1%>%
  filter(species=="RM", treatment=="control", sex=="M")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

rm1=left_join(RM_all_ex, RMex)
rm2=left_join(RM_all_con,RMcon)
rm3=rbind(rm1,rm2)

rm4=rm3%>%
  mutate(proportion=repro/count, species="RM")
write.csv(rm4, "RM_props.csv")
RM_props=read.csv("RM_props.csv")

#combine all species####
```

### Reproductive males dataset  
```{r message=FALSE, error=FALSE}
all_sp=rbind(PB_props, PE_props, PP_props, PF_props, PM_props, RM_props)%>%
  arrange(repro)
head(all_sp)
hist(all_sp$proportion)
```

### Hierarchical beta-binomial model   
#### A. Data Manipulation    
```{r message=FALSE, error=FALSE}
#all_sp$treatment<-ifelse(all_sp$trt=="control", 0, 1) 
#all_sp$years=(all_sp$year-mean(all_sp$year))/(2*sd(all_sp$year)) #standardize year
#all_sp[is.na(all_sp)] <- 0 #set non-detects to 0
#all_sp$spcode=as.integer(all_sp$species) 

#dat_list=list(
# N=length(all_sp$month),
#  y=all_sp$repro,
#  n=all_sp$count,
#  year=all_sp$years,
#  treatment=all_sp$trt,
#  species=all_sp$spcode,
#  month=all_sp$month, 
#  Nmon=length(unique(all_sp$month)),
#  Nsp=length(unique(all_sp$species)))
```
#### B. Model Infrastructure  

```{r message=FALSE, error=FALSE, eval=FALSE}
data{
  int<lower=0> N; // no.of obs
  int <lower=0> y[N];       // reproductive indivs
  int <lower=0>  n[N];       // total males
  vector [N] year;// year
  vector[N] treatment;// treatment
  int month[N]; //ID of each month
  int Nmon; //no.of months
  int species[N]; //species ID
  int Nsp; //no.of species
 }
                
 parameters {
  real alpha;// intercept
  real year_eff; //slope year
  real trt_eff; //slope treatment effect
  real<lower=0> sigma_mon[Nmon];//error for random intercept (month)
  real <lower=0> mon_non;//non-centered error term for species
  real<lower=0> sigma_sp[Nsp];//error for random intercept (species)
  real <lower=0> sp_non;//non-centered error term for month
  real <lower=0> phi;
  real <lower=0, upper=1> pred_repro[N] ;//proportion of reproductive event 
              }
   
  transformed parameters{
  vector <lower=0, upper=1> [N] repro_mu; //mean estimated proportion of reproductive event
  vector <lower=0> [N] A;
  vector <lower=0> [N] B;
  vector [Nmon] alpha_mon; //random intercept per species
   vector [Nsp] alpha_sp; //random intercept per species
  //vector [Nmon] yr_mon; //random slope per month for year effect
  //vector [Nmon] trt_mon;//random slope per month for treatment effect

  
  for (j in 1:Nmon) {
  
  alpha_mon[j]= mon_non*sigma_mon[j];
  }
  
   for (k in 1:Nsp) {
  
  alpha_sp[k]= sp_non*sigma_sp[k];
  }
  
  //model:
  
  for (i in 1:N){
  
  repro_mu[i]= inv_logit(alpha+alpha_mon[month[i]]+alpha_sp[species[i]]+ year_eff*year[i]+trt_eff*treatment[i]);
  }
  
  A = repro_mu * phi;
  B = (1 - repro_mu)* phi;
  
  }
 model {
  //priors
  year_eff~ normal (0,1);
  trt_eff~ normal (0,1);
  mon_non~ normal(0,1);
  sigma_mon~ normal(0,1);
  phi ~normal(0,1);
  sp_non~ normal(0,1);
  sigma_sp~ normal(0,1);
  
  //model likelihood:
  
  pred_repro ~ beta(A, B); // survival estimate, beta dist.
  y~binomial(n, pred_repro); //no.of survivors drawn from binomial dist; based on sample size and reported survival estimate
 
 }
  
  generated quantities {
  
  real pred_y [N];//predictions on proportions
  real log_lik [N];// for looic calculations
  
    pred_y = beta_rng(A, B);
    
    for (x in 1:N){
    log_lik[x]= beta_lpdf(pred_repro[x]| A[x], B[x]);}
   
  }
```

#### C. Model Output  

```{r error=FALSE, message=FALSE}
mod1=readRDS("all_sp_betabinom_mod.RDS")
print(mod1, pars=c("alpha","alpha_mon", "alpha_sp","trt_eff", "year_eff"))
```

#### D. Model Interpretation  

1. community-wide 
* mean community-wide proportion of reproductive males across all years   
```{r error=FALSE, message=FALSE}
#plogis(alpha+mean year_eff+mean trt_eff)
(plogis(-3.08+-0.73+0.25))*100
```

1.a. community-wide (monthly)
* mean community-wide proportion of reproductive males across all years in JAN 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[1]+year_eff+mean trt_eff)
plogis(-3.08+0.44+-0.73+0.25)*100
```

* mean community-wide proportion of reproductive males across all years in FEB 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[2]+year_eff+mean trt_eff)
plogis(-3.08+1.3+-0.73+0.25)*100
```

* mean community-wide proportion of reproductive males across all years in MAR 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[3]+year_eff+mean trt_eff)
plogis(-3.08+2.03+-0.73+0.25)*100
```

* mean community-wide proportion of reproductive males across all years in APR 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[4]+year_eff+mean trt_eff)
plogis(-3.08+2.52+-0.73+0.25)*100
```

* mean community-wide proportion of reproductive males across all years in MAY 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[5]+year_eff+mean trt_eff)
plogis(-3.08+2.32+-0.73+0.25)*100
```

* mean community-wide proportion of reproductive males across all years in JUN 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[6]+year_eff+mean trt_eff)
plogis(-3.08+1.81+-0.73+0.25)*100
```

* mean community-wide proportion of reproductive males across all years in JUL 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[7]+year_eff+mean trt_eff)
plogis(-3.08+1.42+-0.73+0.25)*100
```

* mean community-wide proportion of reproductive males across all years in AUG 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[8]+year_eff+mean trt_eff)
plogis(-3.08+1.3+-0.73+0.25)*100
```

* mean community-wide proportion of reproductive males across all years in SEPT 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[9]+year_eff+mean trt_eff)
plogis(-3.08+1.06+-0.73+0.25)*100
```

* mean community-wide proportion of reproductive males across all years in OCT 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[10]+year_eff+mean trt_eff)
plogis(-3.08+0.73+-0.73+0.25)*100
```

* mean community-wide proportion of reproductive males across all years in NOV 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[11]+year_eff+mean trt_eff)
plogis(-3.08+0.19+-0.73+0.25)*100
```

* mean community-wide proportion of reproductive males across all years in DEC 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[12]+year_eff+mean trt_eff)
plogis(-3.08+0.22+-0.73+0.25)*100
```

2. species-specific  

* PB mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[1]+year_eff+mean trt_eff)
plogis(-3.08+1.20+-0.73+0.25)*100
```

* PE mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[2]+year_eff+mean trt_eff)
plogis(-3.08+0.22+-0.73+0.25)*100
```

* PF mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[3]+year_eff+mean trt_eff)
plogis(-3.08+2.04+-0.73+0.25)*100
```

* PM mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[4]+year_eff+mean trt_eff)
plogis(-3.08+0.99+-0.73+0.25)*100
```

* PP mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[5]+year_eff+mean trt_eff)
plogis(-3.08+1.76+-0.73+0.25)*100
```

* RM mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[6]+year_eff+mean trt_eff)
plogis(-3.08+1.19+-0.73+0.25)*100
```


3. Probability of direction (PD) of effect
```{r error=FALSE, message=FALSE, echo=FALSE}
male_mod=as.data.frame(mod1)
```

* year effect 

```{r error=FALSE, message=FALSE}
length(which(male_mod$year_eff<0))/length(male_mod$year_eff)
```
*There is a >99% probability that year is negatively associated with the number of reproductive male individuals*  

* treatment effect 

```{r error=FALSE, message=FALSE}
length(which(male_mod$trt_eff>0))/length(male_mod$trt_eff)
```
*There is a >99% probability that higher values of set-up(1=exclosure, 0=control) are positively associated with the number of reproductive male individuals*  

#### E. Data Visualization   

1. no.of reproductive individuals across all species in all years per month

```{r error=FALSE, warning=FALSE, message=FALSE}
plot_all_month=all_sp%>%
  ggplot(mapping=aes(x=year, y=repro, fill=trt))+
  geom_col(position = position_dodge())+labs(fill="set-up")+
  xlab("Year")+ylab("No.ofreproductive individuals")+
  facet_wrap(~month)
plot_all_month
```  

2. no. of reproductive males across all years per species per month
```{r error=FALSE, warning=FALSE, message=FALSE}
plot_all_raw=all_sp%>%
  ggplot(mapping=aes(x=reorder(Month,month), y=repro, fill=trt))+
  geom_col(position = position_dodge())+labs(fill="set-up")+
  xlab("Year")+ylab("No.ofreproductive individuals")+
  facet_wrap(~species)
plot_all_raw
```

3. no of reproductive males across all months per year

```{r error=FALSE, warning=FALSE, message=FALSE}
plot_all_year=all_sp%>%
  ggplot(mapping=aes(x=year, y=repro, fill=trt))+
  geom_col(position = position_dodge())+labs(fill="set-up")+
  xlab("Year")+ylab("No.ofreproductive individuals")+
  facet_wrap(~month)
plot_all_year
```  

#### *Questions for Morgan*  

* do the results sort of make sense (year and treatment effect)/ do you agree with how the results were interpreted?  
* I read in the MacLean and Guralnick 2020 paper where they reconstructed breeding phenology of a small mammal that photoperiod was one of the important determinants of their breeding events. would it be worthwhile to add that as a predictor or should we focus on the temporal trends of the proportion of the breeding events and the effect of competition?  
* does it make sense to allow the slope (treatment effect) to vary per month as well? (i.e., does it make sense to expect that at a given month, being in the control set-up might be conducive for them because K-rats may be less active?)  
