---
title: "Phenological traits of Rodents in Portal, AZ (male)"
author: "Pat Dumandan"
date: "11/23/2020"
output: html_document
---

#### Exclosure set-up
```{r, message=FALSE}
library(dplyr)
Portal_rodent=read.csv("./PortalData/Rodents/Portal_rodent.csv")
Portal_plots=read.csv("./PortalData/SiteandMethods/Portal_plots.csv")
portal1=inner_join(Portal_rodent, Portal_plots, by= c("plot", "month", "year"))
```

```{r}
target_sp=c("PB", "PP", "PE", "PF", "PM", "RM")
target_repro=c("S", "M", "R")
target_month=c(1:12)
total_ex=portal1%>%
  filter(species %in% target_sp, testes %in% target_repro,
         treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)
write.csv(total_ex, "male_exclosure_SMR.csv")
```
*if only S in exclosure=1838 (vs 3008 if including M and R)*

##### Control set-up
```{r}
total_con=portal1%>%
  filter(species %in% target_sp, testes %in% target_repro,
         treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)
write.csv(total_con, "male_control_SMR.csv")
```
*if only S in control=1333 (vs 2320 if including M and R)*

## Proportion of counts per month per species
```{r}
#control with scrotal only

#sp1: PB
count_per_month_PBcon= c(4,17,34,19,12,9,7,4,2,5,2,4)
count_per_month_PBex= c(6,14,47,46,34,22,27,10,11,14,3,7)
PBprop=vector(mode="integer", length=length(count_per_month_PBcon))
PBpropex=vector(mode="integer", length=length(count_per_month_PBex))

for (i in 1:length(count_per_month_PBex)) {
  proportions=count_per_month_PBex[i]/sum(count_per_month_PBex)
  PBpropex[i]=proportions
}
PBpropex

for (i in 1:length(count_per_month_PBcon)) {
  proportions=count_per_month_PBcon[i]/sum(count_per_month_PBcon)
  PBprop[i]=proportions
}

PBprop

```

```{r}
# control with S,M,R 
PBcon=portal1%>%
  filter(species=="PB", testes %in% target_repro,
         treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)

PBex=portal1%>%
  filter(species=="PB", testes %in% target_repro,
         treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)

count_per_month_PBcon_smr= c(27,58,95,49,49,35,27,20,21,17,26,21)
count_per_month_PBex_smr= c(43,66,136,95,95,79,73,48,57,47,40,30)
PBprop_smr=vector(mode="integer", length=length(count_per_month_PBcon_smr))
PBpropex_smr=vector(mode="integer", length=length(count_per_month_PBex_smr))

for (i in 1:length(count_per_month_PBcon_smr)) {
  proportions=count_per_month_PBcon_smr[i]/sum(count_per_month_PBcon_smr)
  PBprop_smr[i]=proportions
}
PBprop_smr

for (i in 1:length(count_per_month_PBex_smr)) {
  proportions=count_per_month_PBex_smr[i]/sum(count_per_month_PBex_smr)
  PBpropex_smr[i]=proportions
}
PBpropex_smr
```
*PBcontrol=around 69% of the total "events" occurred in the months of FEB-MAY (~ >10%/month proportions)* 

*PBexclosure=around 72% of the total "events" occurred in the months of MAR-MAY (~ >10%/month proportions)*   

*PB exclosure with all events(SMR), bulk of these events (51%) occurred in MAR-JUN (~ >10%/month proportions)*    

*PB control with all events(SMR), bulk of these events (56%) occurred in FEB-MAY (~ >10%/month proportions)*   

*Qualitatively, it seems that for PBs, the reproductive events (across all years) tend to occur earlier in the control set-ups (Feb-May)*  

### CALCULATE PROPORTION OF EVENTS OCCURRING PER MONTH PER YEAR
#### **to ask Morgan:**   
* is it worth it to add M and R given these counts?  
