---
title: "rodent reproductive phenology (female)"
author: "Pat Dumandan"
date: "01/03/2021"
output: html_document
---

#### A. Data Curation
```{r warning=FALSE, echo=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

Portal_rodent=read.csv("./PortalData/Rodents/Portal_rodent.csv")
Portal_plots=read.csv("./PortalData/SiteandMethods/Portal_plots.csv")
portal1=inner_join(Portal_rodent, Portal_plots, by= c("plot", "month", "year"))%>%
  mutate(Month=recode(month, "1"= "Jan", "2"="Feb", "3"="Mar", "4"="Apr",
                      "5"="May","6"="Jun", "7"="Jul", "8"= "Aug", "9"="Sept",
                      "10"="Oct","11"="Nov", "12"="Dec"))
#female indivs####
target_vag=c("S", "P", "B")
target_nip=c("R", "E", "B")
target_sp=c("PB", "PP", "PE", "PF", "PM", "RM")
portal2=portal1%>%
  filter(vagina %in% target_vag | nipples %in% target_nip |
           pregnant== "P" | lactation=="L")%>%
  filter(species %in%target_sp)

#portal3=portal2%>%
#  filter(species %in%target_sp)

#write.csv(portal3, "female_repro.csv")
#fem_rep=read.csv("female_repro.csv")

#size thresholds####
PBf=portal2%>%
  filter(species=="PB", wgt>=20)%>%
  arrange(wgt)

PPf=portal2%>%
  filter(species=="PP", wgt>=5)%>%
  arrange(wgt)
#have an individual considered juvenile but with enlarged nipples

PEf=portal2%>%
  filter(species=="PE", wgt >=12)%>%
  arrange(wgt)

PFf=portal2%>%
  filter(species=="PF", wgt >=5)%>%
  arrange(wgt)

PMf=portal2%>%
  filter(species=="PM", wgt >=13, treatment=="exclosure")%>%
  arrange(wgt)

RMf=portal2%>%
  filter(species=="RM", wgt >=5)%>%
  arrange(wgt)

#### PMf proportions#####

PMfex=portal2%>%
  filter(species=="PM", wgt >=13, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

PMfcon=portal2%>%
  filter(species=="PM", wgt >=13, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

PMf_all_con=portal1%>%
  filter(species=="PM",sex=="F", treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

PMf_all_ex=portal1%>%
  filter(species=="PM", treatment=="exclosure", sex=="F")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

pmf1=left_join(PMf_all_ex, PMfex)
pmf2=left_join(PMf_all_con,PMfcon)
pmf3=rbind(pmf1,pmf2)

pmf4=pmf3%>%
  mutate(proportion=repro/count, species="PM")
write.csv(pmf4, "PMf_props.csv")
PMf_props=read.csv("PMf_props.csv")

#### PBf proportions#####

PBfex=portal2%>%
  filter(species=="PB", wgt >=20, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

PBfcon=portal2%>%
  filter(species=="PB", wgt >=20, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

PBf_all_con=portal1%>%
  filter(species=="PB",sex=="F", treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

PBf_all_ex=portal1%>%
  filter(species=="PB", treatment=="exclosure", sex=="F")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

pbf1=left_join(PBf_all_ex, PBfex)
pbf2=left_join(PBf_all_con,PBfcon)
pbf3=rbind(pbf1,pbf2)

pbf4=pbf3%>%
  mutate(proportion=repro/count, species="PB")
write.csv(pbf4, "PBf_props.csv")
PBf_props=read.csv("PBf_props.csv")

#### PPf proportions#####

PPfex=portal2%>%
  filter(species=="PP", wgt >=5, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

PPfcon=portal2%>%
  filter(species=="PP", wgt >=5, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

PPf_all_con=portal1%>%
  filter(species=="PP",sex=="F", treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

PPf_all_ex=portal1%>%
  filter(species=="PP", treatment=="exclosure", sex=="F")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

ppf1=left_join(PPf_all_ex, PPfex)
ppf2=left_join(PPf_all_con,PPfcon)
ppf3=rbind(ppf1,ppf2)

ppf4=ppf3%>%
  mutate(proportion=repro/count, species="PP")
write.csv(ppf4, "PPf_props.csv")
PPf_props=read.csv("PPf_props.csv")

#### PEf proportions#####

PEfex=portal2%>%
  filter(species=="PE", wgt >=12, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

PEfcon=portal2%>%
  filter(species=="PE", wgt >=12, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

PEf_all_con=portal1%>%
  filter(species=="PE",sex=="F", treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

PEf_all_ex=portal1%>%
  filter(species=="PE", treatment=="exclosure", sex=="F")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

pef1=left_join(PEf_all_ex, PEfex)
pef2=left_join(PEf_all_con,PEfcon)
pef3=rbind(pef1,pef2)

pef4=pef3%>%
  mutate(proportion=repro/count, species="PE")
write.csv(pef4, "PEf_props.csv")
PEf_props=read.csv("PEf_props.csv")

#### PFf proportions#####

PFfex=portal2%>%
  filter(species=="PF", wgt >=5, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

PFfcon=portal2%>%
  filter(species=="PF", wgt >=5, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

PFf_all_con=portal1%>%
  filter(species=="PF",sex=="F", treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

PFf_all_ex=portal1%>%
  filter(species=="PF", treatment=="exclosure", sex=="F")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

pff1=left_join(PFf_all_ex, PFfex)
pff2=left_join(PFf_all_con,PFfcon)
pff3=rbind(pff1,pff2)

pff4=pff3%>%
  mutate(proportion=repro/count, species="PF")
write.csv(pff4, "PFf_props.csv")
PFf_props=read.csv("PFf_props.csv")

#### RMf proportions#####

RMfex=portal2%>%
  filter(species=="RM", wgt >=5, treatment=="exclosure")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="exclosure")

RMfcon=portal2%>%
  filter(species=="RM", wgt >=5, treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(repro=n())%>%
  mutate(trt="control")

RMf_all_con=portal1%>%
  filter(species=="RM",sex=="F", treatment=="control")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="control")

RMf_all_ex=portal1%>%
  filter(species=="RM", treatment=="exclosure", sex=="F")%>%
  group_by(month,Month, year)%>%
  summarize(count=n())%>%
  mutate(trt="exclosure")

rmf1=left_join(RMf_all_ex, RMfex)
rmf2=left_join(RMf_all_con,RMfcon)
rmf3=rbind(rmf1,rmf2)

rmf4=rmf3%>%
  mutate(proportion=repro/count, species="RM")
write.csv(rmf4, "RMf_props.csv")
RMf_props=read.csv("RMf_props.csv")
```

Reproductive females dataset  
```{r message=FALSE, error=FALSE}
all_female_sp=rbind(PBf_props, PEf_props, PPf_props, PFf_props, PMf_props, RMf_props)%>%
  arrange(repro)
head(all_female_sp)
hist(all_female_sp$proportion)
```
#### B. Data Visualization   

1. No. of reproductive females across all years per species per set-up (how do the numbers of recorded reproductive females change over the years?)
```{r error=FALSE, message=FALSE}
all_female_sp%>%
  ggplot(mapping=aes(x=year, y=repro, fill=trt))+
  geom_col(position=position_dodge())+
  xlab("year")+ylab("No.of reproductive individuals")+
  facet_wrap(~species)+theme(axis.text.x = element_text(angle = 45, hjust=1),
                             panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                             panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  labs(fill="set-up")

```

2. No. of reproductive females across all years per month per set-up per species (how do the numbers of recorded reproductive females change over the course of the year?)
```{r error=FALSE, message=FALSE}
all_female_sp%>%
  ggplot(mapping=aes(x=reorder(Month, month), y=repro, fill=trt))+
  geom_col(position=position_dodge())+
  xlab("year")+ylab("No.of reproductive individuals")+
  facet_wrap(~species)+theme(axis.text.x = element_text(angle = 45, hjust=1),
                             panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                             panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  labs(fill="set-up")

```
3. Total no. of reproductive females across all years per month per set-up (how does the total number of recorded reproductive females change over the course of the year?)
```{r error=FALSE, message=FALSE}
all_female_sp%>%
  ggplot(mapping=aes(x=reorder(Month, month), y=repro, fill=trt))+
  geom_col(position=position_dodge())+
  xlab("year")+ylab("No.of reproductive individuals")+
 theme(axis.text.x = element_text(angle = 45, hjust=1),
                             panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                             panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  labs(fill="set-up")

```
1. Total no. of reproductive females across all years per set-up (how does the total number of recorded reproductive females change over the years?)
```{r error=FALSE, message=FALSE}
all_female_sp%>%
  ggplot(mapping=aes(x=year, y=repro, fill=trt))+
  geom_col(position=position_dodge())+
  xlab("year")+ylab("No.of reproductive individuals")+
  theme(axis.text.x = element_text(angle = 45, hjust=1),
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  labs(fill="set-up")

```

#### C. Model Output  

```{r error=FALSE, message=FALSE}
mod_fem=readRDS("all_female_sp_betabinom_mod.RDS")
print(mod_fem, pars=c("alpha","alpha_mon", "alpha_sp","trt_eff", "year_eff"))
```

#### D. Model Interpretation  

1. community-wide 
* mean community-wide proportion of reproductive females across all years   
```{r error=FALSE, message=FALSE}
#plogis(alpha+mean year_eff+mean trt_eff)
(plogis(-2.58+-0.23+0.16))*100
```

1.a. community-wide (monthly)
* mean community-wide proportion of reproductive females across all years in JAN 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[1]+year_eff+mean trt_eff)
(plogis(-2.58+0.21+-0.23+0.16))*100
```

* mean community-wide proportion of reproductive females across all years in FEB 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[2]+year_eff+mean trt_eff)
(plogis(-2.58+0.60+-0.23+0.16))*100
```

* mean community-wide proportion of reproductive females across all years in MAR 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[3]+year_eff+mean trt_eff)
(plogis(-2.58+1.31+-0.23+0.16))*100
```

* mean community-wide proportion of reproductive females across all years in APR 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[4]+year_eff+mean trt_eff)
(plogis(-2.58+1.87+-0.23+0.16))*100
```

* mean community-wide proportion of reproductive females across all years in MAY 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[5]+year_eff+mean trt_eff)
(plogis(-2.58+2.18+-0.23+0.16))*100
```

* mean community-wide proportion of reproductive females across all years in JUN 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[6]+year_eff+mean trt_eff)
(plogis(-2.58+2.06+-0.23+0.16))*100
```

* mean community-wide proportion of reproductive females across all years in JUL 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[7]+year_eff+mean trt_eff)
(plogis(-2.58+1.51+-0.23+0.16))*100
```

* mean community-wide proportion of reproductive females across all years in AUG 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[8]+year_eff+mean trt_eff)
(plogis(-2.58+1.24+-0.23+0.16))*100
```

* mean community-wide proportion of reproductive females across all years in SEPT 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[9]+year_eff+mean trt_eff)
(plogis(-2.58+0.89+-0.23+0.16))*100
```

* mean community-wide proportion of reproductive females across all years in OCT 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[10]+year_eff+mean trt_eff)
(plogis(-2.58+0.52+-0.23+0.16))*100
```

* mean community-wide proportion of reproductive females across all years in NOV 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[11]+year_eff+mean trt_eff)
(plogis(-2.58+0.28+-0.23+0.16))*100
```

* mean community-wide proportion of reproductive females across all years in DEC 
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[12]+year_eff+mean trt_eff)
(plogis(-2.58+0.18+-0.23+0.16))*100
```

2. species-specific  

* PB mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[1]+year_eff+mean trt_eff)
(plogis(-2.58+0.21+-0.23+0.16))*100
```

* PE mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[2]+year_eff+mean trt_eff)
(plogis(-2.58+1.68+-0.23+0.16))*100
```

* PF mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[3]+year_eff+mean trt_eff)
(plogis(-2.58+0.50+-0.23+0.16))*100
```

* PM mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[4]+year_eff+mean trt_eff)
(plogis(-2.58+0.87+-0.23+0.16))*100
```

* PP mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[5]+year_eff+mean trt_eff)
(plogis(-2.58+1.23+-0.23+0.16))*100
```

* RM mean proportion across all years
```{r error=FALSE, message=FALSE}
#plogis(alpha+alpha[6]+year_eff+mean trt_eff)
plogis(-3.08+1.41+-0.73+0.25)*100
```


3. Probability of direction (PD) of effect
```{r error=FALSE, message=FALSE, echo=FALSE}
female_mod=as.data.frame(mod_fem)
```

* year effect 

```{r error=FALSE, message=FALSE}
length(which(female_mod$year_eff<0))/length(female_mod$year_eff)
```
*There is a >99% probability that year is negatively associated with the number of reproductive male individuals*  

* treatment effect 

```{r error=FALSE, message=FALSE}
length(which(female_mod$trt_eff>0))/length(female_mod$trt_eff)
```
*There is a >99% probability that higher values of set-up(1=exclosure, 0=control) are positively associated with the number of reproductive male individuals*  


#### **Questions for Morgan**  

* do you agree with the selection of the female reproductive traits (all indicated as S,P,B in vagina, R,E,B in nipples, L in lactation, and P in pregnant)?  
* does it make sense to explore the interactive effect of year and treatment?    
* should I be building a model that includes both male and female as predictors? I'm mostly interested in knowing if competition affects the proportion of breeding individuals observed and how it varies over time, and not really if there is a sex-specific signal though.  
* is there reason to believe that breeding cues for females and males would vary among rodents?  (if yes, then maybe I could build a more complex model with an interactive effect of photoperiod or precipitation or something and their sex?)  Not really sure if this idea makes sense  
