---
title: "Reproductive Phenology of PBs"
author: "Pat Dumandan"
date: "2/27/2021"
output: html_document
---

```{r include=FALSE}
require(dplyr)
require(ggplot2)
require(rstan)

options(mc.cores = parallel::detectCores())

tot_dat=read.csv("./reconfigured_data/raw_cleaned/reprod_propn_male.csv")

PB_dat_M=tot_dat%>%filter(species=="PB", !(treatment=="spectabs"))
PB_dat_M[is.na(PB_dat_M)] <- 0 #set non-detects to 0
PB_dat_M$trt<-ifelse(PB_dat_M$treatment=="control", 0, 1) 
PB_dat_M$years=(PB_dat_M$year-mean(PB_dat_M$year))/(2*sd(PB_dat_M$year)) #standardize year
PB_dat_M$months = (PB_dat_M$month - mean(PB_dat_M$month))/(2 *sd(PB_dat_M$month))
PB_dat_M$mon_cos= cos(2*pi*(PB_dat_M$month/12))
PB_dat_M$mon_sin= sin(2*pi*(PB_dat_M$month/12))
PB_dat_M$spcode=as.integer(PB_dat_M$species)

PB_gam_ex=PB_dat_M%>%
  filter(treatment=="exclosure")

require(mgcv)
require(brms)
require(dplyr)
```

## Goal: Identify how the proportion of breeding individuals vary across all months in control vs exclosure plots  
*when do the proportions synchronize and/or diverge at the different treatment plots?*  
*methods employed: GAM and change point model*

```{r}

ggplot(PB_dat_M, aes(y=proportion, x=month, col=treatment)) +
  geom_point() +
  stat_smooth(method = 'gam', formula = y ~ s(x))+
  ggtitle("PB males")+
  scale_x_discrete(name="month", limits=c("Jan", "Feb", "Mar", "Apr","May", "Jun",
                                          "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))+
  theme(axis.text.x= element_text(angle = 90, vjust=0.5, hjust=1),
        panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
```


### **PB control plots**
```{r warning=FALSE, message= FALSE}

PB_gam_con=PB_dat_M%>%
  filter(treatment=="control")%>%
  mutate(date=as.POSIXct(paste(month, "15", year, sep="-"), format="%m-%d-%Y"))
```

**Data Visualization**
```{r message=FALSE}
ggplot(PB_gam_con, aes(y=proportion, x=month)) +
  geom_point() +
  stat_smooth(method = 'gam', formula = y ~ s(x))+
  ggtitle("PB males (control)")+
  scale_x_discrete(name="month", limits=c("Jan", "Feb", "Mar", "Apr","May", "Jun",
                                          "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))+
  theme(axis.text.x= element_text(angle = 90, vjust=0.5, hjust=1),
        panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
```

**GAM with autocorrelation structure**
```{r message=FALSE}
m1=gamm(proportion~s(month, bs="cc", k=12), data=PB_gam_con, correlation = corARMA(form= ~1|year, p=2))
plot(m1$gam)
summary(m1$gam)
```

#### Notes:  
* seems like there is a clear period of high breeding intensity around Feb-Apr in control plots, which coincides with the hot and wet (summer) season? and with the underlying data   
* figure out what the y-lab is in the gam viz  
* figure out what this warning message means: saturated likelihood may be inaccurate when I try family=betar(link="logit"), which also gives negative values for variance explained

**Calculate derivatives: to identify period of change**

```{r warning=FALSE, message= FALSE}
tmpf <- tempfile()
download.file("https://gist.github.com/gavinsimpson/e73f011fdaaab4bb5a30/raw/82118ee30c9ef1254795d2ec6d356a664cc138ab/Deriv.R",tmpf)
source(tmpf)

want=seq(1, nrow(PB_gam_con), length.out = 200)
pdat=with(PB_gam_con, data.frame(year=year[want], month=month[want]))
p2=predict(m1$gam, newdata=pdat, type="terms", se.fit = TRUE)
pdat=transform(pdat, p2=p2$fit[,1], se2=p2$se.fit[,1]) #p2=fit, se2=std.error

df.res=df.residual(m1$gam)
crit.t=qt(0.025, df.res, lower.tail = F)
pdat=transform(pdat, upper=p2+(crit.t*se2), lower=p2-(crit.t*se2))

Term <- "month"
m1.d <- Deriv(m1)
m1.dci <- confint(m1.d, term = Term)
m1.dsig <- signifD(pdat$p2, d = m1.d[[Term]]$deriv,
                     +m1.dci[[Term]]$upper, m1.dci[[Term]]$lower)

plot.Deriv(m1.d, sizer =T)

lines(pdat$month[order(pdat$month)], pdat$p2[order(pdat$month)], 
       xlim=range(pdat$month), ylim=range(pdat$p2), pch=16,
       col="black", lty="dashed")

```

#### Notes:  
* fitted trend line (dashed) vs first derivatives: why do they look different and how do i interpret them? 
* figure out how to extract the values coinciding with the locations where the estimated derivative is positive or negative  


Since we're identifying when rates of change occurred anyway, why not try build a change point model and compare them?  


**Change point model**
```{r message=FALSE, warning=FALSE}
datlist=list(N=length(PB_gam_con$X),year=PB_gam_con$years,
             month=PB_gam_con$month, proportion=PB_gam_con$proportion)


bkpt_con=stan(model_code="
data {
  int<lower=1> N;                      // sample size
  vector<lower=0,upper=1>[N] proportion;        // response 
  vector [N] month;                       // predictor month
  vector [N] year; // predictor year
}

parameters{
  real alpha; // intercept
  real beta1;//slope 1
  real beta2;//slope 2
  real <lower=1, upper=12>bkpt;
  real <lower=0> sigma;
  real year_eff; //year effect
}

transformed parameters{
  real prop_mu [N];

for ( i in 1:N) {
  if (month[i] < bkpt)
    
    prop_mu[i]=alpha+beta1* (month[i]*bkpt)+ year_eff*year[i];
  
  else
    prop_mu[i]= alpha+ beta2 * (month[i]*bkpt+ year_eff*year[i]);
  
}
  }

model {
  
  alpha~normal(0,10);
  beta1~ normal(0,10);
  beta2~normal (0,10);
  bkpt~ uniform(1,12);
  sigma~cauchy(0,2.5);
  year_eff~normal(0,10);
  proportion~normal(prop_mu, sigma); }

generated quantities {
real prop_mu_pred [N];

for (j in 1:N)
prop_mu_pred[j]=normal_rng(prop_mu[j],sigma);

}


", data=datlist, chains=4, iter=1000)
```

Change point model output
```{r message=FALSE}
print(bkpt_con, pars=c("alpha", "beta1", "beta2", "bkpt", "year_eff"))
post=rstan::extract(bkpt_con)$prop_mu
post=as.data.frame(post)
post=t(post)
t1=cbind(PB_gam_con$month,PB_gam_con$year, post)
t1=as.data.frame(t1)
t1=t1%>%
  rename(year=V2, month=V1)
t2=reshape2::melt(t1, id=c("year", "month"))
plot(t2$value~t2$month, type="l", col="grey", ylim=c(0,1))
points(PB_gam_con$proportion~PB_gam_con$month, col="blue", pch=16)

abline(v=2.85, col="black", lty="dashed")
rect(1,0,5,1, col = rgb(0.5,0.5,0.5,1/4))
```
#### Notes:  
* it seems like the model is able to estimate with a bit of high uncertainty when the changepoint is: month 1-5 but mean is around 2-3 (Feb-March) in control plots  
* but what if there are multiple change points? 
* what if I allowed the intercept, slope and breakpoint to vary per treatment group instead?  


### **PB exclosure plots**
```{r warning=FALSE, message= FALSE}
PB_gam_ex=PB_dat_M%>%
  filter(treatment=="exclosure")%>%
  mutate(date=as.POSIXct(paste(month, "15", year, sep="-"), format="%m-%d-%Y"))
```

**Data Visualization**
```{r message=FALSE}
ggplot(PB_gam_ex, aes(y=proportion, x=month)) +
  geom_point() +
  stat_smooth(method = 'gam', formula = y ~ s(x))+
  ggtitle("PB males (exclosure)")+
  scale_x_discrete(name="month", limits=c("Jan", "Feb", "Mar", "Apr","May", "Jun",
                                          "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))+
  theme(axis.text.x= element_text(angle = 90, vjust=0.5, hjust=1),
        panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
```

**GAM with autocorrelation structure**
```{r message=FALSE}
m2=gamm(proportion~s(month, bs="cc", k=12), data=PB_gam_ex, correlation = corARMA(form= ~1|year, p=2))
plot(m2$gam)
summary(m2$gam)
```

**Calculate derivatives: to identify period of change**

```{r warning=FALSE, message= FALSE}

want=seq(1, nrow(PB_gam_ex), length.out = 200)
pdat2=with(PB_gam_ex, data.frame(year=year[want], month=month[want]))
p3=predict(m2$gam, newdata=pdat2, type="terms", se.fit = TRUE)
pdat2=transform(pdat2, p3=p3$fit[,1], se3=p3$se.fit[,1])

df.res=df.residual(m2$gam)
crit.t=qt(0.025, df.res, lower.tail = F)
pdat2=transform(pdat2, upper=p3+(crit.t*se3), lower=p3-(crit.t*se3))

Term <- "month"
m2.d <- Deriv(m2)
m2.dci <- confint(m2.d, term = Term)
m2.dsig <- signifD(pdat2$p3, d = m2.d[[Term]]$deriv,
                     +m2.dci[[Term]]$upper, m2.dci[[Term]]$lower)

plot.Deriv(m2.d, sizer =T)

lines(pdat2$month[order(pdat2$month)], pdat2$p3[order(pdat2$month)], 
       xlim=range(pdat2$month), ylim=range(pdat2$p3), pch=16,
       col="black", lty="dashed")

```

#### Notes:  
*seems like Feb is the inflection period in exclosures but looking at raw data, seems like there isn't a real "peak" period. they just reproduce whenever 

```{r message=FALSE, warning=FALSE}
datlist2=list(N=length(PB_gam_ex$X),year=PB_gam_ex$years,
             month=PB_gam_ex$month, proportion=PB_gam_ex$proportion)


bkpt_ex=stan(model_code="
data {
  int<lower=1> N;                      // sample size
  vector<lower=0,upper=1>[N] proportion;        // response 
  vector [N] month;                       // predictor month
  vector [N] year; // predictor year
}

parameters{
  real alpha; // intercept
  real beta1;//slope 1
  real beta2;//slope 2
  real <lower=1, upper=12>bkpt;
  real <lower=0> sigma;
  real year_eff; //year effect
}

transformed parameters{
  real prop_mu [N];

for ( i in 1:N) {
  if (month[i] < bkpt)
    
    prop_mu[i]=alpha+beta1* (month[i]*bkpt)+ year_eff*year[i];
  
  else
    prop_mu[i]= alpha+ beta2 * (month[i]*bkpt+ year_eff*year[i]);
  
}
  }

model {
  
  alpha~normal(0,10);
  beta1~ normal(0,10);
  beta2~normal (0,10);
  bkpt~ uniform(1,12);
  sigma~cauchy(0,2.5);
  year_eff~normal(0,10);
  proportion~normal(prop_mu, sigma); }

generated quantities {
real prop_mu_pred [N];

for (j in 1:N)
prop_mu_pred[j]=normal_rng(prop_mu[j],sigma);

}


", data=datlist2, chains=4, iter=1000)
```

```{r message=FALSE}
print(bkpt_ex, pars=c("alpha", "beta1", "beta2", "bkpt", "year_eff"))
post1=rstan::extract(bkpt_ex)$prop_mu
post1=as.data.frame(post1)
post1=t(post1)
t3=cbind(PB_gam_ex$month,PB_gam_ex$year, post1)
t3=as.data.frame(t3)
t3=t3%>%
  rename(year=V2, month=V1)
t3=reshape2::melt(t3, id=c("year", "month"))
plot(t3$value~t3$month, type="l", col="grey", ylim=c(0,1))
points(PB_gam_ex$proportion~PB_gam_ex$month, col="blue", pch=16)

abline(v=2, col="black", lty="dashed")
rect(1,0,2,1, col = rgb(0.5,0.5,0.5,1/4))
```
