---
title: "Phenology project"
author: "Pat Dumandan"
---

**I. Data Preparation**  

**Notes:**  
* use Sarah's code to check for bad tags, bad periods and remove them (manually double checked the data on reproductive indivs.as well. Also, with the use of Sarah's code, I have fewer observations now)  
* use only data from 1988-2014 (period when the plot assignments were fairly constant)  
* use only plot-level data from 18 plots (control and exclosure): 1, 2, 4, 8, 9, 11, 12, 14, 17, 22,3, 6, 13, 15, 18, 19, 20, 21  
* proportion of breeders is the number of individuals of a given species and sex in a given plot that exhibits reproductive characteristics divided by the total number of individuals of a given species and sex in a given plot  
* for months when no observations on reproductive individuals were made, set at 0 (not really account for detection bias)  
* lag of 1 set for abiotic covariates  
* all variables except month and year were standardized (centered each value on the mean and divided by two SDs, Gelman 2008). Not really sure how important it is to standardize variables that are treated as smooth terms since I'm not really interested in their effects. I'm just trying to capture the variation they are contributing. 
```{r include=F}
library(dplyr)
library(tidyr)
library(portalr)
library(ggplot2)
library(lubridate)
library(reshape2)
library(splines)
library(rstan)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

source("https://raw.githubusercontent.com/patdumandan/ReproPhenology/main/RScripts/data_cleaning_functions_Supp.R")
```  

```{r warning=FALSE, message=FALSE, echo=FALSE}
####load cleaned data####
Portal_data=summarize_individual_rodents(
  clean = TRUE,
  type = "Rodents",
  length = "all",
  unknowns = FALSE,
  fillweight = FALSE,
  min_plots = 1,
  min_traps = 1,
  download_if_missing = TRUE,
  quiet = FALSE
)%>%filter(!is.na(sex),!(treatment=="spectabs"), !(year<1988), !(year>2014), 
           plot %in%c(1, 2, 4, 8, 9, 11, 12, 14, 17, 22,3, 6, 13, 15, 18, 19, 20, 21))
#Note: 18 plots included based on Ellen's paper

#add note5 column to filter out dead indivs####

Portal_rodent=read.csv("https://raw.githubusercontent.com/weecology/PortalData/main/Rodents/Portal_rodent.csv")
Portal_data_indiv=left_join(Portal_data, Portal_rodent)%>%
  select(period, month, day, year, treatment, plot, stake, species, sex, reprod, age, testes, vagina, pregnant, nipples, lactation,
         hfl, wgt,tag,note2, note5)

#assign tag IDs for untagged individuals (0 and NA in tag column)####

all_tag=id_unknowns(Portal_data_indiv, 19) 

#find and remove bad periods (periods with only one day of trapping)####

Portal_rodent_trapping= read.csv("https://raw.githubusercontent.com/weecology/PortalData/main/Rodents/Portal_rodent_trapping.csv")
tdat=Portal_rodent_trapping%>%group_by(period)%>%summarise(count=sum(sampled))%>%arrange(count)
bad_periods <- filter(tdat, count < 20) #based on Sarah's code
bad_periods <- as.list(bad_periods$period)

Portal_no_badperiod=all_tag[-which(all_tag$period %in%bad_periods),]%>%
  mutate(tag=as.character(tag)) #necessary so function for starred and duplicate tags will not break

#check quality of tags####

#make sure that records with * in note2 are recognized as new individuals

tags = unique(Portal_no_badperiod$tag)
star_tags = starred_tags(dat=Portal_no_badperiod, tags=tags, spp_col=8, tag_col=19)

#locate dead individuals
tags=unique(star_tags$tag)
dead_dat=is_dead(dat=star_tags, tags=tags, spp_col=8, tag_col=19)

# locate records of duplicated tags
tags=unique(dead_dat$tag)
dup_dat= is_duplicate_tag(dat=dead_dat, tags=tags, spp_col=8, tag_col=19) #returns a list of 2
duptags = unique(dup_dat$bad$tag)
no_dup = dup_dat$data[-which(dup_dat$data$tag %in% duptags),] #delete rows flagged as duplicates without clear resolution

# identify records where multiple indivs share same tag in same period
tags = unique(no_dup$tag)
same = same_period(no_dup, tags)

#eliminate tags that appear more than once in the same period - questionable data
sametags = unique(same$tag)
Portal_no_same= no_dup[-which(no_dup$tag %in% sametags),]

# "clean" data

Portal_clean=subsetDat(Portal_no_same)

#Note: this analysis does not necessarily follow the capture history of 
#individuals, what we want are just the event IDs/observations of reprod.
#characteristics to determine peak timing of breeding events

#PB DATASET####

#males####

portal_male=Portal_clean%>%filter(sex=="M", !is.na(sex), !is.na(treatment)) 

repro_male=portal_male%>%
  filter(testes==c("S", "M", "R"))

PB=repro_male%>%
  filter(species=="PB", wgt>=18)

#get count of reproductive males per month per year per trt
pb_dat=PB%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for MALES per month per year per trt
total_PB=portal_male%>%
  filter(species=="PB")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion=right_join(pb_dat, total_PB)%>%
  mutate(proportion=reproductive/abundance, sex="male")%>%
  arrange(proportion)

total_proportion[is.na(total_proportion)] <- 0 #set non-detects to 0

#females####

portal_female=Portal_clean%>%filter(sex=="F") #49% of individuals are males

repro_female=portal_female%>%
  filter(vagina==c("S", "P", "B")| pregnant=="P" | 
           nipples==c("R", "E", "B") | lactation=="L")

PBf=repro_female%>%
  filter(species=="PB", wgt >=21)%>%
  arrange(wgt)

#get count of reproductive males per month per year per trt
pbf_dat=PBf%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for each species per month per year per trt
total_PBf=portal_female%>%
  filter(species=="PB")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion_f=right_join(pbf_dat, total_PBf)%>%
  mutate(proportion=reproductive/abundance, sex="female")%>%
  arrange(proportion)

total_proportion_f[is.na(total_proportion_f)] <- 0 #set non-detects to 0

PB_all=rbind(total_proportion, total_proportion_f)
PB_all=as.data.frame(PB_all)%>%
  mutate(species="PB")

#PP DATASET####

#males####

PP=repro_male%>%
  filter(species=="PP", wgt>=13)%>%
  arrange(wgt)

#get count of reproductive males per month per year per trt
PP_dat=PP%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for MALES per month per year per trt
total_PP=portal_male%>%
  filter(species=="PP")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male were recorded
total_proportion_pp_m=right_join(PP_dat, total_PP)%>%
  mutate(proportion=reproductive/abundance, sex="male")%>%
  arrange(proportion)

total_proportion_pp_m[is.na(total_proportion_pp_m)] <- 0 #set non-detects to 0

#females####

portal_female=Portal_clean%>%filter(sex=="F") #49% of individuals are males

repro_female=portal_female%>%
  filter(vagina==c("S", "P", "B")| pregnant=="P" | nipples==c("R", "E", "B") | lactation=="L")

PPf=repro_female%>%
  filter(species=="PP",  wgt >=12)%>%
  arrange(wgt)

#get count of reproductive males per month per year per trt
PPf_dat=PPf%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for each species per month per year per trt
total_PPf=portal_female%>%
  filter(species=="PP")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion_pp_f=right_join(PPf_dat, total_PPf)%>%
  mutate(proportion=reproductive/abundance, sex="female")%>%
  arrange(proportion)

total_proportion_pp_f[is.na(total_proportion_pp_f)] <- 0 #set non-detects to 0

PP_all=rbind(total_proportion_pp_m, total_proportion_pp_f)
PP_all=as.data.frame(PP_all)%>%
  mutate(species="PP")

#DM DATASET####

#males####

portal_male=Portal_no_badperiod%>%filter(sex=="M") 

repro_male=portal_male%>%
  filter(testes==c("S", "M", "R"))

DM=repro_male%>%
  filter(species=="DM", wgt>=22)

#get count of reproductive males per month per year per trt
DM_dat=DM%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for MALES per month per year per trt
total_DM=portal_male%>%
  filter(species=="DM")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion_dm_m=right_join(DM_dat, total_DM)%>%
  mutate(proportion=reproductive/abundance, sex="male")%>%
  arrange(proportion)

total_proportion_dm_m[is.na(total_proportion_dm_m)] <- 0 #set non-detects to 0

#females####

DMf=repro_female%>%
  filter(species=="DM", wgt>=27)%>%arrange(wgt)

#get count of reproductive males per month per year per trt
DMf_dat=DMf%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for each species per month per year per trt
total_DMf=portal_female%>%
  filter(species=="DM")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion_dm_f=right_join(DMf_dat, total_DMf)%>%
  mutate(proportion=reproductive/abundance, sex="female")%>%
  arrange(proportion)

total_proportion_dm_f[is.na(total_proportion_dm_f)] <- 0 #set non-detects to 0

DM_all=rbind(total_proportion_dm_m, total_proportion_dm_f)
DM_all=as.data.frame(DM_all)%>%
  mutate(species="DM")

#combine PB, PP and DM datasets####
all_sp=rbind(PB_all, PP_all, DM_all)

#add covariates monthly data####

prod=ndvi(level="monthly", sensor="landsat", fill=TRUE)

prod2=prod%>%
  mutate(year=year(date), month=month(date))

#add lag to ndvi####
ndvi_lag=prod2%>%
  mutate(lag_ndvi=lag(ndvi,order_by=date))%>%
  filter(!(year<1988))

temp=weather(level="monthly", fill=TRUE)%>%
  select(year,month,meantemp, mintemp, maxtemp, precipitation, warm_precip, cool_precip)
  
temp$date=as.Date(paste(temp$year, temp$month, 01), "%Y %m %d")

#add lag to weather vars####
temp_lag=temp%>%
  mutate(lag_temp_mean=lag(meantemp,order_by=date),
         lag_temp_min=lag(mintemp,order_by=date),
         lag_temp_max=lag(maxtemp,order_by=date),
         lag_ppt=lag(precipitation,order_by=date),
         lag_ppt_warm=lag(warm_precip,order_by=date),
         lag_ppt_cool=lag(cool_precip,order_by=date))%>%
  filter(!(year<1988))

#combine weather and rodent data####
all_prod=right_join(ndvi_lag,all_sp)
all_prod_temp=right_join(temp_lag, all_prod)

#add biomass data####
bmass=biomass(level="Plot", type="Rodents",
              clean=TRUE, plots="all", time="date", shape="crosstab")%>%
  mutate(month=month(censusdate), date=day(censusdate), year=year(censusdate))%>%
  filter(!(year<1988), !(year>2014), !(treatment %in%c("spectabs", "removal")))

DO_bmass=bmass%>%select(DO, plot, treatment, month, year, date)%>%
  group_by(month, year, treatment)%>%
  summarise(bmass_DO=sum(DO))

DS_bmass=bmass%>%select(DS, plot, treatment, month, year, date)%>%
  group_by(month, year, treatment)%>%
  summarise(bmass_DS=sum(DS))

DM_bmass=bmass%>%select(DM, plot, treatment, month, year, date)%>%
  group_by(month, year, treatment)%>%
  summarise(bmass_DM=sum(DM))

PP_bmass=bmass%>%select(PP, plot, treatment, month, year, date)%>%
  group_by(month, year, treatment)%>%
  summarise(bmass_PP=sum(PP))

PB_bmass=bmass%>%select(PB, plot, treatment, month, year, date)%>%
  group_by(month, year, treatment)%>%
  summarise(bmass_PB=sum(PB))

all_bmass11=left_join(all_prod_temp,DM_bmass, by=c("month", "year", "treatment"))
all_bmass12=left_join(all_bmass11,DO_bmass, by=c("month", "year", "treatment"))
all_bmass13=left_join(all_bmass12,DS_bmass, by=c("month", "year", "treatment"))
all_bmass2=left_join(all_bmass13,PB_bmass, by=c("month", "year", "treatment"))
all_bmass3=left_join(all_bmass2,PP_bmass, by=c("month", "year", "treatment"))%>%
  mutate(DIPO_bmass=rowSums(.[24:26]))%>%drop_na()
all_bmass3$years=(all_bmass3$year-mean(all_bmass3$year))/(2*sd(all_bmass3$year))
all_bmass3$ndvis=(all_bmass3$ndvi-mean(all_bmass3$ndvi))/(sd(all_bmass3$ndvi))
all_bmass3$ndvis_lag=(all_bmass3$lag_ndvi-mean(all_bmass3$lag_ndvi))/(sd(all_bmass3$lag_ndvi))
all_bmass3$temps_lag_mean=(all_bmass3$lag_temp_mean-mean(all_bmass3$lag_temp_mean))/(2*sd(all_bmass3$lag_temp_mean))
all_bmass3$temps_lag_min=(all_bmass3$lag_temp_min-mean(all_bmass3$lag_temp_min))/(2*sd(all_bmass3$lag_temp_min))
all_bmass3$temps_lag_max=(all_bmass3$lag_temp_max-mean(all_bmass3$lag_temp_max))/(2*sd(all_bmass3$lag_temp_max))

#standardize variables####
all_bmass3$temps_mean=(all_bmass3$meantemp-mean(all_bmass3$meantemp))/(2*sd(all_bmass3$meantemp))
all_bmass3$temps_min=(all_bmass3$mintemp-mean(all_bmass3$mintemp))/(2*sd(all_bmass3$mintemp))
all_bmass3$temps_max=(all_bmass3$maxtemp-mean(all_bmass3$maxtemp))/(2*sd(all_bmass3$maxtemp))
all_bmass3$ppts=(all_bmass3$precipitation-mean(all_bmass3$precipitation))/(2*sd(all_bmass3$precipitation))
all_bmass3$ppts_lag=(all_bmass3$lag_ppt-mean(all_bmass3$lag_ppt))/(2*sd(all_bmass3$lag_ppt))
all_bmass3$dipos=(all_bmass3$DIPO_bmass-mean(all_bmass3$DIPO_bmass))/(2*sd(all_bmass3$DIPO_bmass))
all_bmass3$pbs=(all_bmass3$bmass_PB-mean(all_bmass3$bmass_PB))/(2*sd(all_bmass3$bmass_PB))
all_bmass3$pps=(all_bmass3$bmass_PP-mean(all_bmass3$bmass_PP))/(2*sd(all_bmass3$bmass_PP))
all_bmass3$ppts_warm=(all_bmass3$warm_precip-mean(all_bmass3$warm_precip))/(2*sd(all_bmass3$warm_precip))
all_bmass3$ppts_lag_warm=(all_bmass3$lag_ppt_warm-mean(all_bmass3$lag_ppt_warm))/(2*sd(all_bmass3$lag_ppt_warm))
all_bmass3$ppts_lag_cool=(all_bmass3$lag_ppt_cool-mean(all_bmass3$lag_ppt_cool))/(2*sd(all_bmass3$lag_ppt_cool))

#visualization####
pb_plot=all_bmass3%>%filter(species=="PB")
pp_plot=all_bmass3%>%filter(species=="PP")
dm_plot=all_bmass3%>%filter(species=="DM")

```  

```{r echo=F}
head(all_bmass3)
```  


**II. Analyses**  
*Goal*: Reconstruct breeding phenology of male and female PBs, PPs, and DMs in different biotic contexts (control vs k-rat exclosure) and determine the association between breeding odds and abiotic (e.g., NDVI, precipitation, warm_precip, cool_precip, and lags) and biotic (i.e., biomass of PB, PP, and all k-rats(DO,DM,DS)) factors.  

A. Determining which abiotic variables to include in the models and what are the appropriate lags to apply  
*Partial list that Pat thinks should be in the model*

* lag of NDVI (proxy for primary productivity of the system leading to breeding event) 
* lag of mean temperature (potential effect on physiological process--reduced energetic demands for homeothermy at higher temperatures)   
* lag of warm and/or cool precipitation (proxy for primary productivity and also physiological effects for smaller species that may also be inactive during winter)  
*Note: my concern here is that if I add all these, some may be collinear and would be harder to disentangle which one is the most likely driver*  

*Other alternate variables that maybe important*  

* lag of total precipitation (lag of this should be >1 than the lag for NDVI for it to make biological sense in the context of primary productivity)  
* min and max temperature (also with lag of 1; if too low or too high, too early or late for breeding)  
* lag of precipitation but data truncated to when there are no "flooding events" (i.e., when monthly precip > 35 cm or >than average precipitation)  

**Results (exploratory work on abiotic factor effects)**  
In these explorations, I applied GAM approach and assumed linear relationships between the response variable and the abiotic variable  
```{r message=FALSE, include=FALSE}
PB_male_con=pb_plot%>%filter(treatment=="control", sex=="male")
PB_male_ex=pb_plot%>%filter(treatment=="exclosure", sex=="male")
PB_female_con=pb_plot%>%filter(treatment=="control", sex=="female")
PB_female_ex=pb_plot%>%filter(treatment=="exclosure", sex=="female")
```  
1. PB males (w/ lag of NDVI, mean temperature, warm and cool precip)  

```{r message=F}
pbcon_m=mgcv::gam(proportion~s(month, bs="cc")+s(year)+ndvis_lag+ temps_lag_mean+ppts_lag_warm+ppts_lag_cool,
              data=PB_male_con, weights=abundance, family=binomial(link="logit"))
pbex_m=mgcv::gam(proportion~s(month, bs="cc")+s(year)+ndvis_lag+ temps_lag_mean+ppts_lag_warm+ppts_lag_cool,
              data=PB_male_ex, weights=abundance, family=binomial(link="logit"))
```  
a. control  
```{r echo=F}
summary(pbcon_m)
```  
b. exclosure  
```{r echo=F}
summary(pbex_m)
```

2. PB females (w/ lag of NDVI, mean temperature, warm and cool precip)
```{r message=F}
pbcon_f=mgcv::gam(proportion~s(month, bs="cc")+s(year)+ndvis_lag+ temps_lag_mean+ppts_lag_warm+ppts_lag_cool,
              data=PB_female_con, weights=abundance, family=binomial(link="logit"))
pbex_f=mgcv::gam(proportion~s(month, bs="cc")+s(year)+ndvis_lag+ temps_lag_mean+ppts_lag_warm+ppts_lag_cool,
              data=PB_female_ex, weights=abundance, family=binomial(link="logit"))
```  

a. control  
```{r echo=F}
summary(pbcon_f)
```  
b. exclosure  
```{r echo=F}
summary(pbex_f)
```  

```{r message=FALSE, include=FALSE}
PP_male_con=pp_plot%>%filter(treatment=="control", sex=="male")
PP_male_ex=pp_plot%>%filter(treatment=="exclosure", sex=="male")
PP_female_con=pp_plot%>%filter(treatment=="control", sex=="female")
PP_female_ex=pp_plot%>%filter(treatment=="exclosure", sex=="female")
```  
3. PP males (w/ lag of NDVI, mean temperature, warm and cool precip)  
```{r message=F}
ppcon_m=mgcv::gam(proportion~s(month, bs="cc")+s(year)+ndvis_lag+ temps_lag_mean+ppts_lag_warm+ppts_lag_cool,
              data=PP_male_con, weights=abundance, family=binomial(link="logit"))
ppex_m=mgcv::gam(proportion~s(month, bs="cc")+s(year)+ndvis_lag+ temps_lag_mean+ppts_lag_warm+ppts_lag_cool,
              data=PP_male_ex, weights=abundance, family=binomial(link="logit"))
```  

a. control  
```{r echo=F}
summary(ppcon_m)
```  
b. exclosure  
```{r echo=F}
summary(ppex_m)
```  

4. PB females (w/ lag of NDVI, mean temperature, warm and cool precip)
```{r message=F}
ppcon_f=mgcv::gam(proportion~s(month, bs="cc")+s(year)+ndvis_lag+ temps_lag_mean+ppts_lag_warm+ppts_lag_cool,
              data=PP_female_con, weights=abundance, family=binomial(link="logit"))
ppex_f=mgcv::gam(proportion~s(month, bs="cc")+s(year)+ndvis_lag+ temps_lag_mean+ppts_lag_warm+ppts_lag_cool,
              data=PP_female_ex, weights=abundance, family=binomial(link="logit"))
```  
a. control  
```{r echo=F}
summary(ppcon_f)
```  
b. exclosure  
```{r echo=F}
summary(ppex_f)
```  

```{r message=FALSE, include=FALSE}
DM_male_con=dm_plot%>%filter(treatment=="control", sex=="male")
DM_female_con=dm_plot%>%filter(treatment=="control", sex=="female")
```  
3. DM males (w/ lag of NDVI, mean temperature, warm and cool precip)
```{r message=F}
dmcon_m=mgcv::gam(proportion~s(month, bs="cc")+s(year)+ndvis_lag+ temps_mean+ppts_lag_warm+ppts_lag_cool,
              data=DM_male_con, weights=abundance, family=binomial(link="logit"))
dmcon_f=mgcv::gam(proportion~s(month, bs="cc")+s(year)+ndvis_lag+ temps_mean+ppts_lag_warm+ppts_lag_cool,
              data=DM_female_con, weights=abundance, family=binomial(link="logit"))
```  

a. control  
```{r echo=F}
summary(dmcon_m)
```  
b. exclosure  
```{r echo=F}
summary(dmcon_f)
```  

**To discuss with Morgan**  
* In most cases (meaning control and exclosure and either male or female), there was a significant association between P(breeding) and lag of NDVI and precipitation (warm and cool). Lag of NDVI and lag of warm_precip were positively linked to P(breeding), while lag of cool precip was negatively linked to P(breeding) of both PPs and PBs. Does this make sense? Is applying the same lag for NDVI and warm and cool precip appropriate? What maybe biological explanations for these?  
* Contrary to McLean and Guralnick's results, we're not finding strong evidence for the effect of temperature. Is it because of the lag? If yes, should we assume no lag for the effect of temperature? What may be a biological reason to back this assumption? Will rodents decide immediately to breed if the temp.at a given month seems conducive for reproduction? Or should we add a different *type* of temperature like min or max temp. with no lag and assume a negative association between P(breeding) and both types of variables (because either way, breeding earlier or later than is ideal)?  

*Models*:  
* build species- and sex- specific GAMs with increasing levels of complexity (Note: P(breeding)=proportion of individuals exhibiting reproductive characteristics)  
**M1 (abiotic only model)**: P(breeding)~ s(month)+ s(year)+ abiotic variables  
**M2 (abiotic + intraspecific competition)**: P(breeding) ~ s(month)+ s(year)+ abiotic variables + biomass of given population (e.g., if model for PBs, biomass of PBs)  
**M3 (abiotic + intra- and interspecific competition)**: P(breeding) ~ s(month)+ s(year)+ abiotic variables + biomass of given population (e.g., if model for PBs, biomass of PBs) + biomass of other populations   

*Modelling approaches*:  
**mod1**: GAM through mgcv package  
**mod2**: spline model with normal distribution in Stan  
**mod3**: spline model with beta-binomial distribution in Stan  

A. Preliminary Analyses focused on building M1 through different modelling approaches  
*Notes*  
* dataset used is PB females in control     
* for now, using 3 different modelling approaches to build M1 (abiotic only model) to see which one is the most appropriate

First, visualize data so we have a point of reference  (regression line is just proportion~month)

```{r echo=F}
ggplot(PB_female_con, aes(y=proportion, x=month, col=treatment)) +
  geom_point() + 
  ylab("P(breeding)")+
  stat_smooth(method = 'gam', formula = y ~ s(x))+
  ggtitle("PB females in control")+
  scale_x_discrete(name="month", limits=c("Jan", "Feb", "Mar", "Apr","May", "Jun",
                                          "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))+
  theme(axis.text.x= element_text(angle = 90, vjust=0.5, hjust=1),
        panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
```

1. **M1 mod1**  
This model draws from McLean and Guralnick's paper (2020) where they built a separate GAM to reconstruct breeding phenology of PEMA and then make separate GLMMs to test for the relationship of breeding odds with breeding cues (i.e., photoperiod, temperature, precipitation). Not sure why they didn't just make a single GAM and set the environmental variables as linear terms. Here, we use a logistic function and cubic cyclic regression spline smoothers (bs="cc" because breeding events are cyclical).  

```{r message=FALSE}
PB_female_con=pb_plot%>%filter(treatment=="control", sex=="female")
#no explicit biotic effects

mod1=mgcv::gam(proportion~s(month, bs="cc")+s(year)+ ndvis_lag+ temps_lag_mean+ppts_lag_warm+ppts_lag_cool, data=PB_female_con, method = 'REML', weights = abundance, family = binomial)
```  

```{r echo=FALSE}
summary(mod1) 
```  

```{r echo=F, fig.cap="Fig.1. Partial effects plot (mod1)"}
plot(mod1, shade=T, page=1)
```  

2. **M1 mod2**  
This model is my attempt to make a simple spline model in Stan regardless of distribution (assumed to be normal). Here, I directly use the proportion data so not really weighting them properly. I am not really able to account for potential overdispersion and variation in the sample sizes (e.g., 1 reproductive in 2 total individuals is treated similarly as 2 reproductive in 4 individuals). Here, we use the "splines" package to create the B-splines with degree set at 3 (cubic cyclic) outside Stan and then add it as "data". I could probably build the b-splines directly into Stan but this seems easier, and would make the model less complex. Plus, I still have other variables to add so making it less complex at this stage may be wise.  

```{r message=F, warning=F}  
X1 =PB_female_con$month
X2= PB_female_con$year
X3= PB_female_con$ndvis_lag
X4= PB_female_con$temps_lag_mean
X5= PB_female_con$ppts_lag_warm
X6= PB_female_con$ppts_lag_cool
num_data = length(X1)
B1 = t(bs(X1, df=NULL, knots=NULL, degree=3, intercept = FALSE)) # creating the B-splines for month, degree=3 for cubic spline
B2 = t(bs(X2, df=NULL, knots=NULL, degree=3, intercept = FALSE)) #creating B-splines for year
num_basis1 = nrow(B1)
num_basis2 = nrow(B2)
Y = PB_female_con$proportion

dat_list=list(X1 =PB_female_con$month,
              X2= PB_female_con$year,
              X3= PB_female_con$ndvis_lag,
              X4= PB_female_con$temps_lag_mean,
              X5= PB_female_con$ppts_lag_warm,
              X6= PB_female_con$ppts_lag_cool,
              num_data = length(X1),
              B1 = t(bs(X1, df=NULL, knots=NULL, degree=3, intercept = FALSE)), 
              B2 = t(bs(X2, df=NULL, knots=NULL, degree=3, intercept = FALSE)),
              num_basis1 = nrow(B1),
              num_basis2 = nrow(B2),
              Y = PB_female_con$proportion)

mod2<-stan(model_code="
data { 
  int num_data; //rows of observations 
  int num_basis1; //no. of basis (order-1) for month 
  int num_basis2; //no. of basis (order-1) for year
  real <lower=0,upper=1>Y[num_data]; //response variable (e.g., no.of breeding obs.)
  vector[num_data] X3; //lag_NDVI
  vector[num_data] X4; //lag_temp
  vector[num_data] X5;//lag warm precip
  vector[num_data] X6;//lag cool precip
  matrix[num_basis1, num_data] B1; //matrix of coefficients of splines(rows), length of X1 (columns)
   matrix[num_basis2, num_data] B2; //matrix of coefficients of splines(rows), length of X2 (columns)
} 
 
parameters { 
  row_vector[num_basis1] a_raw; // global smooth term for month
  row_vector[num_basis2] b_raw; // global smooth term for year
  real a0; //intercept
  real ndvi_eff;
  real temp_eff;
  real warm_ppt_eff;
  real cool_ppt_eff;
  
  real<lower=0> sigma; //error term
  real<lower=0> tau; // for noncentered parameterization of spline coefficients (month)
  real<lower=0> phi; // for noncentered parameterization of spline coefficients (year)
  
} 
 
transformed parameters { 
  row_vector[num_basis1] a; //coefficient for month smooth term
  row_vector[num_basis2] b; //coefficient for year smooth term
  
  vector[num_data] Y_hat; 
  
  a = a_raw*tau;  
  b = b_raw*phi;
  
 Y_hat=a0 + ndvi_eff*X3 +temp_eff*X4+ warm_ppt_eff*X5+ cool_ppt_eff*X6 + to_vector(a*B1)+ to_vector(b*B2);
 }

model { 
  a_raw ~ normal(0, 1); 
  b_raw ~ normal(0, 1); 
  phi~normal(0,1);
  tau ~ normal(0, 1); 
  sigma ~ normal(0, 1); 
  ndvi_eff~normal(0,1);
  temp_eff~normal(0,1);
  warm_ppt_eff~normal(0,1);
  cool_ppt_eff~normal(0,1);
  Y~ normal(Y_hat, sigma);//likelihood
  }",iter=300,
data =dat_list, chains=2)
```  

```{r echo=F}
print(mod2, pars=c("a0", "ndvi_eff", "warm_ppt_eff", "cool_ppt_eff", "temp_eff"))
```  
*Notes:*  
* Intercept way off (ra data suggests 8%). for this it's ~50%, overestimation. Also, the effects of the different covariates don't really make sense.  Although, the posterior draws are somewhat within the range of the expected.  

```{r echo=FALSE, fig.cap="Fig.2a. spline fit over raw data (mod2)"}
#plotting splines####
ff<-extract(mod2)
Y_hat_med <- array(NA, length(Y)) #median estimate
Y_hat_ub <- array(NA, length(Y)) #upper boundary
Y_hat_lb <- array(NA, length(Y)) #lower boundary

for (i in 1:length(Y)) {
  Y_hat_med[i] <- median(ff$Y_hat[,i]);
  Y_hat_lb[i] <- quantile(ff$Y_hat[,i],probs = 0.025)
  Y_hat_ub[i] <- quantile(ff$Y_hat[,i],probs = 0.975)
}

plot(X1,Y, xaxt="n", ylab="P(breeding)", xlab="month") #plot raw data
axis(1, c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"), at=c(1:12))
lines(smooth.spline(X1, Y_hat_med), col="blue")
lines(smooth.spline(X1, Y_hat_lb), lty=2, col="red") #0.025
lines(smooth.spline(X1, Y_hat_ub), lty=2, col="red") #0.975
```  

```{r echo=FALSE, fig.cap="Fig.2b. posterior draws over raw data (mod2)"}

yrep=extract(mod2)$Y
Y_hat_med <- array(NA, length(Y)) #median estimate
for (i in 1:length(Y)) {
  Y_hat_med[i] <- mean(ff$Y_hat[,i])}

#extract posterior draws####
post1=rstan::extract(mod2)$Y
post1=as.data.frame(post1)
post1=t(post1)
t3=cbind(PB_female_con$month, post1)
t3=as.data.frame(t3)
t3=t3%>%
  rename("month"="V1")
t3=reshape2::melt(t3, id=c("month"))

#plot posterior draws####
plot(t3$value~t3$month, type="l", col="grey", ylim=c(0,1), xaxt="n",  ylab="P(breeding)", xlab="month")
axis(1, c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"), at=c(1:12))
points(PB_female_con$proportion~PB_female_con$month, col="blue", pch=16)
lines(smooth.spline(X1, Y_hat_med), col="red")

```

3. **M1 mod3**  
This model has added complexity by assuming a beta-binomial (BB) distribution. This may be appropriate because if we use a binomial distribution, the probabilities are fixed for a certain number of trials (i.e., total indivs). But for BB model, the probabilities vary from trial to trial (which we may expect from month to month). This will help us account for the variation in sample sizes. Also, the BB model has an overdispersion term that may be appropriate especially for PPs that are less active/reproductive during winter, and will produce a lot of 0s.  

$$ 
 \begin{align*}
 
 Y_{ij} \sim Binom \ (n_{ij}, \ p_{ij}) \qquad \text{(Eqn.1)} \\
 p_i \sim Beta \ (\gamma_1, \gamma_2)\qquad \text{(Eqn.2)} \\
\gamma_1 = \mu_1 * \phi ; \quad \gamma_2 = (1 - \mu_1) * \phi \qquad \text { (Eqn.3)} \\
\mu_1 = logit^{-1} (\alpha + \Sigma (\beta_i*X_i)+ \Sigma (f_1(Y_i))) \qquad \text { (Eqn. 4)}

\end{align*}
$$  

In Eqn. 1, Y is the number of individuals exhibiting reproductive characteristics for species *i* in plot *j* out of *n* individuals of the same species in the same plot, with an underlying probability (*p*) of being reproductive. The *p* of an individual to exhibit reproductive characteristics is drawn from a beta distribution with shape parameters (Eqn.2). The shape parameter \(\gamma_1\) is a function of the mean *p* of being in reproductive state \(\mu\) multiplied by an overdispersion parameter \(\phi\), and the shape parameter \(\gamma_2\) is the product of \(\phi\) and \(\gamma_1\). Finally, the mean *p* of being in reproductive state is the logistic function of the global intercept \(\alpha\), the sum of the effects of the linear terms \(\Sigma (\beta_i*X_i)\) and the effects of the non-linear terms \(\Sigma (f_1(Y_i))\).  

*Note*: The sum of the linear terms will vary per level of complexity (if it is abiotic only, intra- or intra-and inter) but the sum of the linear terms will be the same all throughout with smoothed terms for month and year.  

```{r message=F, warning=F}  

X1= PB_female_con$month
X2= PB_female_con$year
X3= PB_female_con$ndvis_lag
X4= PB_female_con$temps_lag_mean
X5= PB_female_con$ppts_lag_warm
X6= PB_female_con$ppts_lag_cool
num_data= length(X1)
B1= t(bs(X1, df=NULL, knots=NULL, degree=3, intercept = FALSE)) 
B2= t(bs(X2, df=NULL, knots=NULL, degree=3, intercept = FALSE))
num_basis1= nrow(B1)
num_basis2= nrow(B2)
Y= PB_female_con$reproductive
n= PB_female_con$abundance

dat_list2=list(X1= PB_female_con$month,
               X2= PB_female_con$year,
               X3= PB_female_con$ndvis_lag,
               X4= PB_female_con$temps_lag_mean,
               X5= PB_female_con$ppts_lag_warm,
               X6= PB_female_con$ppts_lag_cool,
               num_data = length(X1),
               B1 = t(bs(X1, df=NULL, knots=NULL, degree=3, intercept = FALSE)), 
               B2 = t(bs(X2, df=NULL, knots=NULL, degree=3, intercept = FALSE)),
               num_basis1 = nrow(B1),
               num_basis2 = nrow(B2),
               Y = PB_female_con$reproductive,
               n=PB_female_con$abundance)

mod3<-stan(model_code="
data { 
  int num_data; //rows of observations 
  int num_basis1; //no. of basis (order-1) 
  int num_basis2; //no. of basis (order-1)
  
  int Y[num_data]; //response variable (# of reproductive)
  int n[num_data]; //total no.of indivs in plot
   
  vector[num_data] X3; //lag_NDVI
  vector[num_data] X4; //lag_temp
  vector[num_data] X5; //lag warm precip
  vector[num_data] X6;// lag cool precip
  
  matrix[num_basis1, num_data] B1; //matrix of coefficients of splines(rows), length of X1 (columns)
  matrix[num_basis2, num_data] B2; //matrix of coefficients of splines(rows), length of X2 (columns)

} 
 
parameters { 
 row_vector[num_basis1] a_raw; // smooth terms for month
  row_vector[num_basis2] b_raw; // smooth terms for year
  real a0; //intercept
  real ndvi_eff;
  real cool_ppt_eff;
  real temp_eff;
  real warm_ppt_eff;
  
  real<lower=0> sigma; //error term for shape params
  real<lower=0> tau; // for noncentered parameterization of spline coefficients (month)
 // real<lower=0> phi; // for noncentered parameterization of spline coefficients (year)
  
 vector <lower=0, upper=1> [num_data] pred_repro;//breeding odds as a parameter
} 
 
transformed parameters { 
  row_vector[num_basis1] a; //noncentered parameters of splines
  row_vector[num_basis2] b; //noncentered parameters of splines
  
  //beta dist of probability as a deterministic function
  vector <lower=0, upper=1> [num_data] Y_hat; // mean of response variable
 
 //beta shape params
  vector <lower=0> [num_data] a1;
  vector <lower=0> [num_data] b1;
  
  a = a_raw*tau;  
  b = b_raw*tau; 
  
  Y_hat = inv_logit(a0 + ndvi_eff*X3 +temp_eff*X4+ warm_ppt_eff*X5+ cool_ppt_eff*X6 + to_vector(a*B1)+ to_vector(b*B2)); 
  

a1=Y_hat*sigma;
b1=(1-Y_hat)*sigma;
}

model { 
  a0~ normal(0,1);
  a_raw ~ normal(0, 1); 
  b_raw ~ normal(0, 1);
  tau ~ normal(0, 1); 
//  phi ~ normal(0, 1);
  sigma ~ normal(0, 1); 
  ndvi_eff~normal(0,1);
  warm_ppt_eff~normal(0,1);
  cool_ppt_eff~normal(0,1);
  temp_eff~normal(0,1);
  pred_repro ~ beta(a1, b1);
  Y~ binomial(n, pred_repro);
}",
iter=2000, control=list(max_treedepth=12, adapt_delta=0.99),
data =dat_list2)
```  

```{r echo=F}
print(mod3, pars=c ("a0", "ndvi_eff", "warm_ppt_eff", "cool_ppt_eff", "temp_eff"))
```  
*Notes:*  
* The intercept is still off. I'm not sure why. But the effect estimates for the different abiotic covariates now makes sense, negative for cool_precip, positive for NDVI, warm_precip and temperature. So, yay!  There is high uncertainty in the spline fit (Fig.3a) though.And I'm not sure why the posterior draws look very erratic.  

```{r echo=F, fig.cap="Fig.3a. spline fit over raw data (mod3)"}

#plotting regression lines over raw data####
ff<-extract(mod3)
Y_hat_med <- array(NA, length(Y)) #median estimate
Y_hat_ub <- array(NA, length(Y)) #upper boundary
Y_hat_lb <- array(NA, length(Y)) #lower boundary

for (i in 1:length(Y)) {
  Y_hat_med[i] <- median(ff$pred_repro[,i]);
  Y_hat_lb[i] <- quantile(ff$pred_repro[,i],probs = 0.025)
  Y_hat_ub[i] <- quantile(ff$pred_repro[,i],probs = 0.975)
}

prop=PB_female_con$proportion
plot(X1,prop, xaxt="n", ylab="P(breeding)", xlab="month") #plot raw data
axis(1, c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"), at=c(1:12))
lines(smooth.spline(X1, Y_hat_med), col="blue")
lines(smooth.spline(X1, Y_hat_ub), lty=2, col="red")
lines(smooth.spline(X1, Y_hat_lb), lty=2, col="red")
```  

```{r echo=F, fig.cap="Fig.3b. posterior draws over raw data (mod3)"}
yrep=extract(mod3)$pred_repro
Y_hat_med <- array(NA, length(Y)) #median estimate
for (i in 1:length(Y)) {
  Y_hat_med[i] <- mean(ff$pred_repro[,i])}

#plot posterior draws####
post1=rstan::extract(mod3)$pred_repro
post1=as.data.frame(post1)
post1=t(post1)
t3=cbind(PB_female_con$month, post1)
t3=as.data.frame(t3)
t3=t3%>%
  rename("month"="V1")
t3=reshape2::melt(t3, id=c("month"))

#plot posterior draws####
plot(t3$value~t3$month, type="l", col="grey", ylim=c(0,1), xaxt="n", ylab="P(breeding)", xlab="month")
axis(1, c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"), at=c(1:12))
points(PB_female_ex$proportion~PB_female_ex$month, col="blue", pch=16)
lines(smooth.spline(X1, Y_hat_med), col="red")
```  

**To figure out/ discuss with Chris**  
* how to calculate derivatives from Stan output (in generated quantities block? maybe check Gavin's gist for creating Deriv function?)    
* how problematic is 1 divergent transition (below diagonal plot)  
* do I need to standardize the predictors for the smooth terms? or does it not matter? 
* how important is it to add a smoothing prior?  
* will I get the same results if I use a binomial distribution with a logit parameterization rather than a beta-binomial model?  
