---
title: "PB prelim analysis"
author: "Pat Dumandan"
---

```{r, setup, include=FALSE}
require(knotr)
knitr::opts_chunk$set(echo = TRUE)
getwd()
```

### **Data Curation**
```{r warning=FALSE, message=FALSE}
require(portalr)
require(dplyr)
require(ggplot2)
require (lme4)
require(rstan)
require(brms)
require(rstanarm)
require(bayesplot)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

####load cleaned individual-level data####
Portal_data_indiv=summarize_individual_rodents(
  path = get_default_data_path(),
  clean = TRUE,
  type = "Rodents",
  length = "all",
  unknowns = FALSE,
  time = "date",
  fillweight = FALSE,
  min_plots = 1,
  min_traps = 1,
  download_if_missing = TRUE,
  quiet = FALSE
)%>%filter(!(treatment=="removal")& !is.na(treatment)& !is.na(sex))%>%
  mutate(month=as.character(month))

# add columns for reproductive traits
Portal_data=load_rodent_data(clean = TRUE)
Portal_rodent=Portal_data[["rodent_data"]]%>%mutate(month=factor(month))%>%
  filter(!is.na(sex))

#create full dataset without removal plots
portal1=left_join(Portal_data_indiv, Portal_rodent)%>%
  filter(!(treatment=="removal")& !is.na(treatment)& !is.na(sex))%>%
  mutate(month=as.character(month), Month=recode(month, "1"= "Jan", "2"="Feb", "3"="Mar", "4"="Apr",
                                           "5"="May","6"="Jun", "7"="Jul", "8"="Aug", "9"="Sept",
                                           "10"="Oct","11"="Nov", "12"="Dec"))%>%
  select(period, month, Month, day, year, plot, stake,
         treatment, species, sex, reprod, vagina, nipples,lactation, pregnant, testes,hfl,wgt, tag)

str(portal1)
```

### **Data Manipulation** 

```{r warning=FALSE, message=FALSE}
portal_male=portal1%>%filter(sex=="M") #49% of individuals are males
head(portal_male)

####determine threshold for breeding adult male individuals####

target_repro=c("S", "M", "R")
repro_male=portal_male%>%
  filter(testes==c("S", "M", "R"))
head(repro_male)

#BA=repro_male%>%
#  filter(species=="BA")%>%
#  arrange(wgt)

#size thresholds####
BA=repro_male%>%
  filter(species=="BA", wgt >=6)

DM=repro_male%>%
  filter(species=="DM", wgt >=15)

DO=repro_male%>%
  filter(species=="DO", wgt >=29)

DS=repro_male%>%
  filter(species=="DS", wgt >=12)

NEA=repro_male%>%
  filter(species=="NA", wgt >=121)

OL=repro_male%>%
  filter(species=="OL", wgt >=19)

OT=repro_male%>%
  filter(species=="OT", wgt >=10)

PH=repro_male%>%
  filter(species=="PH", wgt >=18)

PL=repro_male%>%
  filter(species=="PL", wgt >=20)

PB=repro_male%>%
  filter(species=="PB", wgt >=16)

PP=repro_male%>%
  filter(species=="PP", wgt >=10)

PE=repro_male%>%
  filter(species=="PE", wgt >=7)

PI=repro_male%>%
  filter(species=="PI", wgt >=15)

PF=repro_male%>%
  filter(species=="PF", wgt >=4)

PM=repro_male%>%
  filter(species=="PM", wgt >=11)

RM=repro_male%>%
  filter(species=="RM", wgt >=4)

RF=repro_male%>%
  filter(species=="RF", wgt >=11)

RO=repro_male%>%
  filter(species=="RO", wgt >=6)

SF=repro_male%>%
  filter(species=="SF", wgt >=39)

SH=repro_male%>%
  filter(species=="SH", wgt >=51)

SO=repro_male%>%
  filter(species=="SO", wgt >=68)

#create full dataset with all species

full_repro_male_dat=rbind(SO,SH,SF,RO,RM,RF,PP,PM,PL,PH,PF,PE,PB,OT,OL,NEA,DS,DO,DM,BA, PI)
full_repro_male_dat=as.data.frame(full_repro_male_dat)
head(full_repro_male_dat)

####calculate proportion of reproductive individuals####

#get count of reproductive males for each species per month per year per trt
repro_dat=full_repro_male_dat%>%
  group_by(month, year, treatment, species)%>%
  summarise(reproductive=n())

#get total observed abundance for each species per month per year per trt
total_rodents=portal_male%>%
  group_by(month,year, treatment, species)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion=right_join(repro_dat, total_rodents)%>%
  mutate(proportion=reproductive/abundance)%>%
  arrange(proportion)
head(total_proportion)
#length(unique(total_proportion$species)) #21 spp
#max(total_proportion$proportion, na.rm=T) #1
```

### **Data Visualization** 
```{r warning=FALSE, message=FALSE, echo=FALSE}
#sample using PB data only
PB_dat_M=total_proportion%>%filter(species=="PB", !(treatment=="spectabs"))
plot(PB_dat_M$proportion~PB_dat_M$year)
####plot PB proportion across all years per trt type
PB_dat_M%>%
  ggplot(mapping = aes(x=year, y=proportion, colour=treatment))+
  geom_point()+geom_smooth(se = FALSE, method = 'loess')

####plot PB proportion across all years per trt type per month
PB_dat_M%>%
  ggplot(mapping = aes(x=year, y=proportion, colour=treatment))+
  geom_point()+geom_smooth(se = FALSE, method = 'lm')+
  facet_wrap(~month)
PB_dat_M=as.data.frame(PB_dat_M)
####plot PB proportion across all years per month per trt type
#control
PB_dat_M_con=total_proportion%>%filter(species=="PB", treatment=="control")
PB_dat_M_con%>%
  ggplot(mapping = aes(x=year, y=proportion))+
  geom_point()+geom_smooth(se = FALSE, method = 'lm')+
  facet_wrap(~month)+ggtitle("PB male control")

#exclosure
PB_dat_M_ex=total_proportion%>%filter(species=="PB", treatment=="exclosure")
PB_dat_M_ex%>%
  ggplot(mapping = aes(x=year, y=proportion))+
  geom_point()+geom_smooth(se = FALSE, method = 'lm')+
  facet_wrap(~month)+ggtitle("PB male exclosure")

```

### Data Analysis  

#### simple binomial regression with logit link (intercept only)  

```{r warning=FALSE, message=FALSE, echo=FALSE}
mod_int=glm(proportion~1, data=PB_dat_M, family=binomial(link="logit"))
summary(mod_int) #AIC: 163.72
```  

#### binomial regression with year and trt
```{r warning=FALSE, message=FALSE, echo=FALSE}
PB_dat_M$trt<-ifelse(PB_dat_M$treatment=="control", 0, 1) 
PB_dat_M$years=(PB_dat_M$year-mean(PB_dat_M$year))/(2*sd(PB_dat_M$year)) #standardize year

mod_int2=glm(proportion~1+years+trt, data=PB_dat_M, family=binomial(link="logit"))
summary(mod_int2) #AIC: 159.94
```  

#### binomial regression with year and trt and month as circular variable
```{r warning=FALSE, message=FALSE, echo=FALSE}
PB_dat_M$trt<-ifelse(PB_dat_M$treatment=="control", 0, 1) 
PB_dat_M$years=(PB_dat_M$year-mean(PB_dat_M$year))/(2*sd(PB_dat_M$year)) #standardize year
PB_dat_M$month=as.integer(PB_dat_M$month)
mod_int3=glm(proportion~1+years+trt+cospi(month)+sinpi(month), data=PB_dat_M, family=binomial(link="logit"))
summary(mod_int3) #AIC: 161.95
```  

#### hierarchical binomial regression with year and trt as covar and month as random intercept
```{r warning=FALSE, message=FALSE, echo=FALSE}
PB_dat_M$trt<-ifelse(PB_dat_M$treatment=="control", 0, 1) 
PB_dat_M$years=(PB_dat_M$year-mean(PB_dat_M$year))/(2*sd(PB_dat_M$year)) #standardize year
PB_dat_M$month=as.integer(PB_dat_M$month)
mod_int4=glmer(proportion~1+years+trt+(1|month), data=PB_dat_M, family=binomial(link="logit"))
summary(mod_int4) #AIC: 101.8
```  

*problem with binomial is: assumption that the probability of the number of reproductive individuals(proportion) is constant*  
*beta-binomial allows us to model the proportion/probability of reproductive observations as a free parameter*

#### simple beta-binomial (intercept only)
```{r warning=FALSE, message=FALSE}
mod_int5=readRDS("~/ReproPhenology/model_output/PB_intercept.RDS")
print(mod_int5, pars=c("alpha", "phi"))

#posterior predictive checks
y=PB_dat_M$proportion
yrep=extract(mod_int5$pred_y)
ppc_overlay(y, yrep)

#mod_int5=stan(model_code="
#   data{
#  int<lower=0> N; // no.of obs
#  int <lower=0> y[N];       // reproductive indivs
#  int <lower=0>  n[N];       // total males
# }
                
# parameters {
#  real alpha;// intercept
#  real <lower=0> phi;
#  real <lower=0, upper=1> pred_repro[N] ;//proportion of reproductive indivs 
#              }
   
 # transformed parameters{
#  vector <lower=0, upper=1> [N] repro_mu; //so we can add statement describing proportion (not able to do in parameters block)
#  vector <lower=0> [N] A;
#  vector <lower=0> [N] B;

#  //model:
  
#  for (i in 1:N){
  
#  repro_mu[i]= inv_logit(alpha);
#  }
  
#  A = repro_mu * phi;
#  B = (1 - repro_mu)* phi;
#}

# model {
#  //priors
# alpha~normal(0,1);  
# phi ~normal(0,1);
  
#  //model likelihood:
  
#  pred_repro ~ beta(A, B); // survival estimate, beta dist.
#  y~binomial(n, pred_repro); //no.of survivors drawn from binomial dist; based on sample size and reported survival estimate
 
# }
  
#  generated quantities {
  
# real pred_y [N];//predictions on proportions
#  real log_lik [N];// for looic calculations
  
 #   pred_y = beta_rng(A, B);
    
#    for (x in 1:N){
#    log_lik[x]= beta_lpdf(pred_repro[x]| A[x], B[x]);}
#   }    ", data=dat_list, chains=4, iter=3000)
```
  
*model performance: loo(m1_loo) #-562.9 +/- 26.2*  

#### beta-binomial with year and treatment as predictors

```{r warning=FALSE, message=FALSE}
PB_intyrtrt=readRDS("./model_output/PB_intyrtrt.RDS")
print(PB_intyrtrt, pars=c("alpha", "phi", "year_eff", "trt_eff"))
# parameters {
#  real alpha;// intercept
#  real year_eff; //slope year
#  real trt_eff; //slope treatment effect
#  real <lower=0> phi;
# real <lower=0, upper=1> pred_repro[N] ;//proportion of reproductive event 
#              }

# transformed parameters {
#  //model:
  
#  for (i in 1:N){
  
#  repro_mu[i]= inv_logit(alpha+ year_eff*year[i]+trt_eff*treatment[i]);
#  }}

```
*model performance: loo(m3_loo) #-592.4+27.8*  

### beta-binomial with year, treatment, and trigonometric functions to account for periodicity(month)  
*trigonometri
```{r warning=FALSE, message=FALSE}
#mon_cos= cos(2*pi*(month/12))
#sin_cos= sin(2*pi*(month/12))
 # model
#  for (i in 1:N){
  
#  repro_mu[i]= inv_logit(alpha+ year_eff*year[i]+trt_eff*treatment[i]+monc_eff*mon_cos[i]+mons_eff*mon_sin[i]);
#  }

PB_intyrtrtmon=readRDS("./model_output/PB_intyrtrtmon.RDS")
```
loo(m3_loo) #-592.3+27.7