---
title: "Phenological traits"
author: "Pat Dumandan"
date: "11/03/2020"
output: html_document
---

```{r, warning=FALSE,message=FALSE}
library(devtools)
library(dplyr)
library(phest)
library(ggplot2)
library(date)
#install_github("willpearse/phest")

Portal_rodent=read.csv("./PortalData/Rodents/Portal_rodent.csv")
Portal_plots=read.csv("./PortalData/SiteandMethods/Portal_plots.csv")
portal1=inner_join(Portal_rodent, Portal_plots, by= c("plot", "month", "year"))
#head(portal1)
```

#### PB control and exclosure
```{r}
PBmaleex=portal1%>%
  filter(species=="PB", testes=="S",treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)

#convert to Julian dates
PBdate=as.data.frame(mdy.date(PBmaleex$month, PBmaleex$day, PBmaleex$year))
PBmaleex1=cbind(PBmaleex, PBdate)
PBmaleex1$Julian=PBmaleex1$`mdy.date(PBmaleex$month, PBmaleex$day, PBmaleex$year)`

PBmalecon=portal1%>%
  filter(species=="PB", testes=="S",treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
PBcdate=as.data.frame(mdy.date(PBmalecon$month, PBmalecon$day, PBmalecon$year))
PBmalecon1=cbind(PBmalecon, PBcdate)
PBmalecon1$Julian=PBmalecon1$`mdy.date(PBmalecon$month, PBmalecon$day, PBmalecon$year)`

#estimate onset
phest::weib.limit(PBmalecon1$Julian)

date.mdy(13500) 

```

#### PP Male control and exclosure

```{r}
PPmaleex=portal1%>%
  filter(species=="PP", testes=="R",treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
PPdate=as.data.frame(mdy.date(PPmaleex$month, PPmaleex$day, PPmaleex$year))
PPmaleex1=cbind(PPmaleex, PPdate)
PPmaleex1$Julian=PPmaleex1$`mdy.date(PPmaleex$month, PPmaleex$day, PPmaleex$year)`

#estimate onset
phest::weib.limit(PPmaleex1$Julian, upper=T)
date.mdy(21765)

PPmalecon=portal1%>%
  filter(species=="PP", testes=="R",treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
PPcdate=as.data.frame(mdy.date(PPmalecon$month, PPmalecon$day, PPmalecon$year))
PPmalecon1=cbind(PPmalecon, PPcdate)
PPmalecon1$Julian=PPmalecon1$`mdy.date(PPmalecon$month, PPmalecon$day, PPmalecon$year)`

#estimate onset
phest::weib.limit(PPmalecon1$Julian, upper=T)

date.mdy(21765)  

```

#### PF male control and exclosure

```{r}
PFmaleex=portal1%>%
  filter(species=="PF", testes=="M",treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
PFdate=as.data.frame(mdy.date(PFmaleex$month, PFmaleex$day, PFmaleex$year))
PFmaleex1=cbind(PFmaleex, PFdate)
PFmaleex1$Julian=PFmaleex1$`mdy.date(PFmaleex$month, PFmaleex$day, PFmaleex$year)`

#estimate onset
phest::weib.limit(PFmaleex1$Julian, upper=T)
date.mdy(22798)

PFmalecon=portal1%>%
  filter(species=="PF", testes=="R",treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
PFcdate=as.data.frame(mdy.date(PFmalecon$month, PFmalecon$day, PFmalecon$year))
PFmalecon1=cbind(PFmalecon, PFcdate)
PFmalecon1$Julian=PFmalecon1$`mdy.date(PFmalecon$month, PFmalecon$day, PFmalecon$year)`

#estimate onset
phest::weib.limit(PFmalecon1$Julian, upper=T)

date.mdy(21292)  

```

#### PE male control and exclosure

```{r}
PEmaleex=portal1%>%
  filter(species=="PE", testes=="R",treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
PEdate=as.data.frame(mdy.date(PEmaleex$month, PEmaleex$day, PEmaleex$year))
PEmaleex1=cbind(PEmaleex, PEdate)
PEmaleex1$Julian=PEmaleex1$`mdy.date(PEmaleex$month, PEmaleex$day, PEmaleex$year)`

#estimate onset
phest::weib.limit(PEmaleex1$Julian, upper=T)
#THIS GIVES ME AN ESTIMATE NOT JUST UPPER AND LOWER CIs! but why??

date.mdy(23589)

PEmalecon=portal1%>%
  filter(species=="PE", testes=="R",treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
PEcdate=as.data.frame(mdy.date(PEmalecon$month, PEmalecon$day, PEmalecon$year))
PEmalecon1=cbind(PEmalecon, PEcdate)
PEmalecon1$Julian=PEmalecon1$`mdy.date(PEmalecon$month, PEmalecon$day, PEmalecon$year)`

#estimate onset
phest::weib.limit(PEmalecon1$Julian, upper=T)

date.mdy(20826)  
```

#### PM male control and exclosure

```{r}
PMmaleex=portal1%>%
  filter(species=="PM", testes=="R",treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
PMdate=as.data.frame(mdy.date(PMmaleex$month, PMmaleex$day, PMmaleex$year))
PMmaleex1=cbind(PMmaleex, PMdate)
PMmaleex1$Julian=PMmaleex1$`mdy.date(PMmaleex$month, PMmaleex$day, PMmaleex$year)`

#estimate onset
phest::weib.limit(PMmaleex1$Julian, upper=T)
#THIS GIVES ME AN ESTIMATE NOT JUST UPPER AND LOWER CIs! but why??

date.mdy(21654.13)

PMmalecon=portal1%>%
  filter(species=="PM", testes=="R",treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
PMcdate=as.data.frame(mdy.date(PMmalecon$month, PMmalecon$day, PMmalecon$year))
PMmalecon1=cbind(PMmalecon, PMcdate)
PMmalecon1$Julian=PMmalecon1$`mdy.date(PMmalecon$month, PMmalecon$day, PMmalecon$year)`

#estimate onset
phest::weib.limit(PMmalecon1$Julian, upper=T)

date.mdy(7764)  
```

#### RM male control and exclosure

``` {r}
RMmaleex=portal1%>%
  filter(species=="RM", testes=="R",treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
RMdate=as.data.frame(mdy.date(RMmaleex$month, RMmaleex$day, RMmaleex$year))
RMmaleex1=cbind(RMmaleex, RMdate)
RMmaleex1$Julian=RMmaleex1$`mdy.date(RMmaleex$month, RMmaleex$day, RMmaleex$year)`

#estimate onset
phest::weib.limit(RMmaleex1$Julian, upper=T)
#THIS GIVES ME AN ESTIMATE NOT JUST UPPER AND LOWER CIs! but why??

date.mdy(21832)

RMmalecon=portal1%>%
  filter(species=="RM", testes=="R",treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
RMcdate=as.data.frame(mdy.date(RMmalecon$month, RMmalecon$day, RMmalecon$year))
RMmalecon1=cbind(RMmalecon, RMcdate)
RMmalecon1$Julian=RMmalecon1$`mdy.date(RMmalecon$month, RMmalecon$day, RMmalecon$year)`

#estimate onset
phest::weib.limit(RMmalecon1$Julian, upper=T)

date.mdy(21330)  

```

#### DM male control and exclosure
Ooohh, I didn't know K-rats werent **totally** excluded in exclosures
**breeding season for Dipodomys in California: late winter to early spring (Kenagy, 1985)**
```{r}
DMmaleex=portal1%>%
  filter(species=="DM", testes=="R",treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
DMdate=as.data.frame(mdy.date(DMmaleex$month, DMmaleex$day, DMmaleex$year))
DMmaleex1=cbind(DMmaleex, DMdate)
DMmaleex1$Julian=DMmaleex1$`mdy.date(DMmaleex$month, DMmaleex$day, DMmaleex$year)`

#estimate onset
phest::weib.limit(DMmaleex1$Julian) #, upper=T)
date.mdy(6495)

DMmalecon=portal1%>%
  filter(species=="DM", testes=="R",treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
DMdate=as.data.frame(mdy.date(DMmalecon$month, DMmalecon$day, DMmalecon$year))
DMmalecon1=cbind(DMmalecon, DMdate)
DMmalecon1$Julian=DMmalecon1$`mdy.date(DMmalecon$month, DMmalecon$day, DMmalecon$year)`

#estimate onset
phest::weib.limit(DMmaleex1$Julian, upper=T)
date.mdy(22725)
#warning message:Repeated earliest measurements; applying correction

#count how many per month
jan=DMmalecon%>%
    filter(month==9)
```

#### DO male control and exclosure
**breeding season Feb-Jun (Johnston, 1956)**

```{r}
DOmaleex=portal1%>%
  filter(species=="DO", testes=="R",treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
DOdate=as.data.frame(mdy.date(DOmaleex$month, DOmaleex$day, DOmaleex$year))
DOmaleex1=cbind(DOmaleex, DOdate)
DOmaleex1$Julian=DOmaleex1$`mdy.date(DOmaleex$month, DOmaleex$day, DOmaleex$year)`

#estimate onset
phest::weib.limit(DOmaleex1$Julian, upper=T)
date.mdy(21618)

DOmalecon=portal1%>%
  filter(species=="DO", testes=="R",treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species)

#convert to Julian dates
DOdate=as.data.frame(mdy.date(DOmalecon$month, DOmalecon$day, DOmalecon$year))
DOmalecon1=cbind(DOmalecon, DOdate)
DOmalecon1$Julian=DOmalecon1$`mdy.date(DOmalecon$month, DOmalecon$day, DOmalecon$year)`

#estimate onset
phest::weib.limit(DOmalecon1$Julian, upper=T)
date.mdy(22027)
#warning message:Repeated earliest measurements; applying correction

#count how many per month
jan=DOmaleex%>%
    filter(month==9)

```
#### DS male control and exclosure

```{r}
DSmaleex=portal1%>%
  filter(species=="DS", testes=="R",treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species)
```
**Maybe not include? too few samples (only 6 events)**

## Plots for repro. timing

```{r}
maletraits=read.csv("male_S_month.csv")

DMmaletraits=maletraits%>%
  filter(Species=="DM")

DMmaletraits%>%
  ggplot(data=DMmaletraits, mapping=aes(x=reorder(Month, Month_num), y=Count, fill=Treatment))+
  geom_col(position=position_dodge())+
  xlab("Month")+ylab("No.of event observations")

PBmaletraits=maletraits%>%
  filter(Species=="PB")

PBmaletraits%>%
  ggplot(data=PBmaletraits, mapping=aes(x=reorder(Month, Month_num), y=Count, fill=Treatment))+
  geom_col(position=position_dodge())+
  xlab("Month")+ylab("No.of event observations")

DOmaletraits=maletraits%>%
  filter(Species=="DO")

DOmaletraits%>%
  ggplot(data=DOmaletraits, mapping=aes(x=reorder(Month, Month_num), y=Count, fill=Treatment))+
  geom_col(position=position_dodge())+
  xlab("Month")+ylab("No.of event observations")

PPmaletraits=maletraits%>%
  filter(Species=="PP")

PPmaletraits%>%
  ggplot(data=PPmaletraits, mapping=aes(x=reorder(Month, Month_num), y=Count, fill=Treatment))+
  geom_col(position=position_dodge())+
  xlab("Month")+ylab("No.of event observations")

PFmaletraits=maletraits%>%
  filter(Species=="PF")

PFmaletraits%>%
  ggplot(data=PFmaletraits, mapping=aes(x=reorder(Month, Month_num), y=Count, fill=Treatment))+
  geom_col(position=position_dodge())+
  xlab("Month")+ylab("No.of event observations")

PEmaletraits=maletraits%>%
  filter(Species=="PE")

PEmaletraits%>%
  ggplot(data=PEmaletraits, mapping=aes(x=reorder(Month, Month_num), y=Count, fill=Treatment))+
  geom_col(position=position_dodge())+
  xlab("Month")+ylab("No.of event observations")

PMmaletraits=maletraits%>%
  filter(Species=="PM")

PMmaletraits%>%
  ggplot(data=PMmaletraits, mapping=aes(x=reorder(Month, Month_num), y=Count, fill=Treatment))+
  geom_col(position=position_dodge())+
  xlab("Month")+ylab("No.of event observations")

RMmaletraits=maletraits%>%
  filter(Species=="RM")

RMmaletraits%>%
  ggplot(data=RMmaletraits, mapping=aes(x=reorder(Month, Month_num), y=Count, fill=Treatment))+
  geom_col(position=position_dodge())+
  xlab("Month")+ylab("No.of event observations")

``` 

## *TO FIGURE OUT/ TO DISCUSS WITH MORGAN:*  
* why I'm not getting median estimates for the onset for some species (sample size?)  (maybe my input data is weird??)
* how to get estimates that does not include year (but Julian date...)  
* should I include DS (but too few samples in exclosure <10)
* Why are there K-rat species in exclosures? how should I account for that?  
* if i want to see if the onset of events is moving earlier/later, should I just focus on identifying the first observation date per month per year?  
* should I choose the months with greatest proportion of reproductive trait observations, choose when the earliest is for that period and then collect dates annually?  
* will it be interesting to know the proportion of individuals that exhibit reproductive characteristics annually, and see if there is a pattern in control and exclosure set-ups, and/or see if there are fewer/more of them being reproductively prepared? 