---
title: "Phenology project_update(v4)"
author: "Pat Dumandan"
date: "4/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(portalr)
library(ggplot2)
library(lubridate)

tmpf <- tempfile()
download.file("https://gist.github.com/gavinsimpson/e73f011fdaaab4bb5a30/raw/82118ee30c9ef1254795d2ec6d356a664cc138ab/Deriv.R",tmpf)
source(tmpf)
```

``` {r warning=FALSE, message=FALSE, include=FALSE}
#LOAD INDIVIDUAL LEVEL-DATA
Portal_data_indiv=summarize_individual_rodents(
  clean = TRUE,
  type = "Rodents",
  length = "all",
  unknowns = FALSE,
  time = "date",
  fillweight = FALSE,
  min_plots = 1,
  min_traps = 1,
  download_if_missing = TRUE,
  quiet = FALSE
)%>%filter(!is.na(sex),!(treatment=="spectabs"), !(year<1988), !(year>2014), plot %in%c(1, 2, 4, 8, 9, 11, 12, 14, 17, 22,3, 6, 13, 15, 18, 19, 20, 21))

#PB DATASET####

#males####

portal_male=Portal_data_indiv%>%filter(sex=="M") 
head(portal_male)

repro_male=portal_male%>%
  filter(testes==c("S", "M", "R"))

PB=repro_male%>%
  filter(species=="PB", wgt>=16)

#get count of reproductive males per month per year per trt
pb_dat=PB%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for MALES per month per year per trt
total_PB=portal_male%>%
  filter(species=="PB")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion_pb_m=right_join(pb_dat, total_PB)%>%
  mutate(proportion=reproductive/abundance, sex="male")%>%
  arrange(proportion)

total_proportion_pb_m[is.na(total_proportion_pb_m)] <- 0 #set non-detects to 0

#females####

portal_female=Portal_data_indiv%>%filter(sex=="F") #49% of individuals are males
head(portal_male)

repro_female=portal_female%>%
  filter(vagina==c("S", "P", "B")| pregnant=="P" | nipples==c("R", "E", "B") | lactation=="L")

PBf=repro_female%>%
  filter(species=="PB", wgt >=21)

#get count of reproductive males per month per year per trt
pbf_dat=PBf%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for each species per month per year per trt
total_PBf=portal_female%>%
  filter(species=="PB")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion_pb_f=right_join(pbf_dat, total_PBf)%>%
  mutate(proportion=reproductive/abundance, sex="female")%>%
  arrange(proportion)

total_proportion_pb_f[is.na(total_proportion_pb_f)] <- 0 #set non-detects to 0

#PP DATASET####

#males####
PP=repro_male%>%
  filter(species=="PP", wgt >=10)

#get count of reproductive males per month per year per trt
PP_dat=PP%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for MALES per month per year per trt
total_PP=portal_male%>%
  filter(species=="PP")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion_pp_m=right_join(PP_dat, total_PP)%>%
  mutate(proportion=reproductive/abundance, sex="male")%>%
  arrange(proportion)

total_proportion_pp_m[is.na(total_proportion_pp_m)] <- 0 #set non-detects to 0

#females####

PPf=repro_female%>%
  filter(species=="PP", wgt >=12)

#get count of reproductive males per month per year per trt
PPf_dat=PPf%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for each species per month per year per trt
total_PPf=portal_female%>%
  filter(species=="PP")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion_pp_f=right_join(PPf_dat, total_PPf)%>%
  mutate(proportion=reproductive/abundance, sex="female")%>%
  arrange(proportion)

total_proportion_pp_f[is.na(total_proportion_pp_f)] <- 0 #set non-detects to 0

#DM DATASET####

#males####

DM=repro_male%>%
  filter(species=="DM", wgt>=22)

#get count of reproductive males per month per year per trt
DM_dat=DM%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for MALES per month per year per trt
total_DM=portal_male%>%
  filter(species=="DM")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion_dm_m=right_join(DM_dat, total_DM)%>%
  mutate(proportion=reproductive/abundance, sex="male")%>%
  arrange(proportion)

total_proportion_dm_m[is.na(total_proportion_dm_m)] <- 0 #set non-detects to 0

#females####

DMf=repro_female%>%
  filter(species=="DM", wgt >=27)

#get count of reproductive males per month per year per trt
DMf_dat=DMf%>%
  group_by(month, year, treatment)%>%
  summarise(reproductive=n())

#get total observed abundance for each species per month per year per trt
total_DMf=portal_female%>%
  filter(species=="DM")%>%
  group_by(month,year, treatment)%>%
  summarise(abundance=n())

#calculate proportion
#this creates NAs for months when no reproductive male was recorded
total_proportion_dm_f=right_join(DMf_dat, total_DMf)%>%
  mutate(proportion=reproductive/abundance, sex="female")%>%
  arrange(proportion)

total_proportion_dm_f[is.na(total_proportion_dm_f)] <- 0 #set non-detects to 0

#COMBINE SEX-SPECIFIC DATA FOR EACH SPECIES####
PB_all=rbind(total_proportion_pb_m, total_proportion_pb_f)
PB_all=as.data.frame(PB_all)%>%
  mutate(species="PB")

PP_all=rbind(total_proportion_pp_m, total_proportion_pp_f)
PP_all=as.data.frame(PP_all)%>%
  mutate(species="PP")

DM_all=rbind(total_proportion_dm_m, total_proportion_dm_f)
DM_all=as.data.frame(DM_all)%>%
  mutate(species="DM")

#combine PB, PP and DM datasets####
all_sp=rbind(PB_all, PP_all, DM_all)

#add ndvi and precip monthly data####

prod=ndvi(level="monthly", sensor="landsat", fill=TRUE)

prod2=prod%>%
  mutate(year=year(date), month=month(date))

ppt=weather(level="monthly", fill=TRUE)%>%select(year,month,precipitation)

all_prod=right_join(prod2,all_sp)
all_prod_ppt=right_join(ppt, all_prod)%>%filter(!is.na(precipitation), !is.na(ndvi))

#add biomass data####
bmass=biomass(path="repo", level="Plot", type="Rodents",
              clean=TRUE, plots="all", time="date", shape="crosstab")%>%
  mutate(month=month(censusdate), date=day(censusdate), year=year(censusdate))%>%
   filter(!(year<1988), !(year>2014), !(treatment=="spectabs"), plot%in%c(1, 2, 4, 8, 9, 11, 12, 14, 17, 22,3, 6, 13, 15, 18, 19, 20, 21))

DM_bmass=bmass%>%select(DM, plot, treatment, censusdate)%>%
  mutate(month=month(censusdate), date=day(censusdate), year=year(censusdate))%>%
  group_by(month, year, treatment)%>%
  summarise(bmass_DM=sum(DM))

PB_bmass=bmass%>%select(PB, plot, treatment, censusdate)%>%
  mutate(month=month(censusdate), date=day(censusdate), year=year(censusdate))%>%
  group_by(month, year, treatment)%>%
  summarise(bmass_PB=sum(PB))

PP_bmass=bmass%>%select(PP, plot, treatment, censusdate)%>%
  mutate(month=month(censusdate), date=day(censusdate), year=year(censusdate))%>%
  group_by(month, year, treatment)%>%
  summarise(bmass_PP=sum(PP))

all_bmass=left_join(all_prod_ppt, DM_bmass, by=c("month", "year", "treatment"))
all2_bmass=left_join(all_bmass, PB_bmass, by=c("month", "year", "treatment"))
all3_bmass=left_join(all2_bmass, PP_bmass, by=c("month", "year", "treatment"))%>%
  filter(!is.na(bmass_DM), !is.na(bmass_PB), !is.na(bmass_PP))

pb_plot=all3_bmass%>%filter(species=="PB")
pp_plot=all3_bmass%>%filter(species=="PP")
dm_plot=all3_bmass%>%filter(species=="DM")
```

```{r include=FALSE}
#1. proportion
#total breeders/total abundance

#2. proportion of males and females in control vs exclosure
#total repro M (or F) in control(or exclosure)/total M (or F) in control and exclosure


#3. proportion of males vs females
# total repro M(or F) in control (or exclosure)/ total repro M and F in control (or exclosure)

```  

### PB MODELS 

```{r message=FALSE, include=FALSE}
PB_male_con=pb_plot%>%filter(treatment=="control", sex=="male")
PB_male_ex=pb_plot%>%filter(treatment=="exclosure", sex=="male")
PB_female_con=pb_plot%>%filter(treatment=="control", sex=="female")
PB_female_ex=pb_plot%>%filter(treatment=="exclosure", sex=="female")
```


##### PB MALES IN CONTROL
```{r message=FALSE, echo=FALSE}
#no explicit biotic effects
m1=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+lag_ndvi+lag_ppt, data=wt, method = 'REML', weights = abundance, family = binomial)

summary(m1) #R2=28.1%

#add effect of DM biomass
m1_bmass=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+ndvi+precipitation+bmass_PB
                   , data=var_lag2, method = 'REML', weights = abundance, family = binomial)

summary(m1_bmass) #R2=31.7%

#add effect of DM biomass AND intraspecific competition (PB)
m1_bmass2=mgcv::gam(proportion~s(month.y, bs="cc", k=12)+s(year)+ndvi+precipitation+bmass_DM+
                   bmass_PB+bmass_PP, data=var_lag2, method = 'REML', weights = abundance, family = binomial)

summary(m1_bmass2) #R2=35.7%

```

```{r message=FALSE}
#best fit plot####
plot(m1_bmass2, shade=T, pages=1)
```  

Fig.1. Partial effects plot of the relationships between different predictor variables and the proportion of breeding male PBs in control plots estimated using a generalized additive model (GAM).  

```{r message=FALSE, echo=FALSE, warning=FALSE}
want=seq(1, nrow(PB_male_con), length.out = 200)
pdat2=with(PB_male_con, data.frame(year=year[want], month=month[want], ndvi=ndvi[want], 
                                   precipitation=precipitation[want], bmass_DM=bmass_DM[want],
                                   bmass_PB=bmass_PB[want], bmass_PP=bmass_PP[want]))
p3=predict(m1_bmass2, newdata=pdat2, type="terms", se.fit = TRUE)
pdat2=transform(pdat2, p3=p3$fit[,1], se3=p3$se.fit[,1]) #p2=fit, se2=std.error

df.res=df.residual(m1_bmass)

crit.t=qt(0.025, df.res, lower.tail = F)
pdat2=transform(pdat2, upper=p3+(crit.t*se3), lower=p3-(crit.t*se3))

m1.d <- Deriv(m1_bmass2)
Term="month"
m1.dci <- confint(m1.d, term = Term)
m1.dsig <- signifD(pdat2$p3, d = m1.d[[Term]]$deriv,
                   +m1.dci[[Term]]$upper, m1.dci[[Term]]$lower)
plot.Deriv(m1.d, sizer=T, term=Term, xaxt="n")
axis(1, at=1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
abline(v=3.1, type="l", lty=2)

```  

Fig.2. Periods of significant change in the seasonal trends in the reproductive phenology of breeding male Bailey's pocket mouse (*Chaetodipus baileyi*) in control plots in a project site near Portal, Arizona. Periods of significant increase (blue) or decrease (red) are indicated by the coloured sections of the trend. Dashed vertical line indicate peak breeding period.


#### PB MALES IN EXCLOSURE  

```{r message=FALSE, echo=FALSE}

#exclosure####
#no explicit biotic effects
m2=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)
             , data=PB_male_ex, method = 'REML', weights = abundance, family = binomial)

summary(m2) #R2=35.5%

#add effect of DM biomass
m2_bmass=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+s(bmass_PB)
                   , data=PB_male_ex, method = 'REML', weights = abundance, family = binomial)

summary(m2_bmass) #R2=37%

#add effect of DM biomass AND intraspecific competition (PB competition)
m2_bmass2=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+
                      s(bmass_PB)+s(bmass_PP), data=PB_male_ex, method = 'REML', weights = abundance, family = binomial)

summary(m2_bmass2) #R2=44.9%
``` 

```{r message=FALSE}

#best fit plot####
plot(m2_bmass2, shade=T, pages=1)

```  

Fig.3. Partial effects plot of the relationships between different predictor variables and the proportion of breeding male PBs in k-rat exclosure plots estimated using a generalized additive model (GAM).   

**FOR DISCUSSION WITH MORGAN**: In choosing this specific "best fit" model, I chose the one that had fewer predictors even though the % deviance explained isn't necessarily the highest because the additional predictor didn't necessarily have any strong significant relationships.  


```{r message=FALSE, echo=FALSE, warning=FALSE}
want=seq(1, nrow(PB_male_ex), length.out = 200)
pdat2=with(PB_male_ex, data.frame(year=year[want], month=month[want], ndvi=ndvi[want], 
                                   precipitation=precipitation[want], bmass_DM=bmass_DM[want],
                                   bmass_PB=bmass_PB[want], bmass_PP=bmass_PP[want]))
p3=predict(m2_bmass2, newdata=pdat2, type="terms", se.fit = TRUE)
pdat2=transform(pdat2, p3=p3$fit[,1], se3=p3$se.fit[,1]) #p2=fit, se2=std.error

df.res=df.residual(m2_bmass2)

crit.t=qt(0.025, df.res, lower.tail = F)
pdat2=transform(pdat2, upper=p3+(crit.t*se3), lower=p3-(crit.t*se3))

m2.d <- Deriv(m2_bmass2)
Term="month"
m2.dci <- confint(m2.d, term = Term)
m2.dsig <- signifD(pdat2$p3, d = m2.d[[Term]]$deriv,
                   +m2.dci[[Term]]$upper, m2.dci[[Term]]$lower)
plot.Deriv(m2.d, sizer=T, term=Term, xaxt="n")
axis(1, at=1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
abline(v=2.7, type="l", lty=2)
```  

Fig.4. Periods of significant change in the seasonal trends in the reproductive phenology of breeding male Bailey's pocket mouse (*Chaetodipus baileyi*) in k-rat exclosure plots in a project site near Portal, Arizona. Periods of significant increase (blue) or decrease (red) are indicated by the coloured sections of the trend. Dashed vertical line indicate peak breeding period.  

#### PB FEMALES IN CONTROL 

```{r message=FALSE, echo=FALSE}
#no explicit biotic effects
m3=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)
             , data=PB_female_con, method = 'REML', weights = abundance, family = binomial)

summary(m3) #R2=41.6%

#add effect of DM biomass
m3_bmass=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+s(bmass_PB)
                   , data=PB_female_con, method = 'REML', weights = abundance, family = binomial)

summary(m3_bmass) #R2=50.9%

#add effect of DM biomass AND intraspecific competition (PB competition)
m3_bmass2=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+
                      s(bmass_PB)+s(bmass_PP), data=PB_female_con, method = 'REML', weights = abundance, family = binomial)

summary(m3_bmass2) #R2=58.9%
```  

```{r message=FALSE}

#best fit plot####
plot(m3_bmass2, shade=T, pages=1)
```  

Fig.5. Partial effects plot of the relationships between different predictor variables and the proportion of breeding female PBs in control plots estimated using a generalized additive model (GAM).  

```{r message=FALSE, warning-FALSE, echo=FALSE}
want=seq(1, nrow(PB_female_con), length.out = 200)
pdat2=with(PB_female_con, data.frame(year=year[want], month=month[want], ndvi=ndvi[want], 
                                  precipitation=precipitation[want], bmass_DM=bmass_DM[want],
                                  bmass_PB=bmass_PB[want], bmass_PP=bmass_PP[want]))
p3=predict(m3_bmass2, newdata=pdat2, type="terms", se.fit = TRUE)
pdat2=transform(pdat2, p3=p3$fit[,1], se3=p3$se.fit[,1]) #p2=fit, se2=std.error

df.res=df.residual(m3_bmass)

crit.t=qt(0.025, df.res, lower.tail = F)
pdat2=transform(pdat2, upper=p3+(crit.t*se3), lower=p3-(crit.t*se3))

m3.d <- Deriv(m3_bmass2)
Term="month"
m3.dci <- confint(m3.d, term = Term)
m3.dsig <- signifD(pdat2$p3, d = m3.d[[Term]]$deriv,
                   +m3.dci[[Term]]$upper, m3.dci[[Term]]$lower)
plot.Deriv(m3.d, sizer=T, term=Term, xaxt="n")
axis(1, at=1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
abline(v=4.2, type="l", lty=2)

```  

Fig.6. Periods of significant change in the seasonal trends in the reproductive phenology of breeding female Bailey's pocket mouse (*Chaetodipus baileyi*) in control plots in a project site near Portal, Arizona. Periods of significant increase (blue) or decrease (red) are indicated by the coloured sections of the trend. Dashed vertical line indicate peak breeding period.  

#### PB FEMALES IN EXCLOSURE  

```{r message=FALSE, echo=FALSE}
#no explicit biotic effects
m4=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)
             , data=PB_female_ex, method = 'REML', weights = abundance, family = binomial)

summary(m4) #R2=53.5%

#add effect of DM biomass
m4_bmass=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+s(bmass_PB)
                   , data=PB_female_ex, method = 'REML', weights = abundance, family = binomial)

summary(m4_bmass) #R2=55%

#add effect of DM biomass AND intraspecific competition (PB competition)
m4_bmass2=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+
                      s(bmass_PB)+s(bmass_PP), data=PB_female_ex, method = 'REML', weights = abundance, family = binomial)

summary(m4_bmass2) #R2=68.3%
```  

```{r message=FALSE, echo=FALSE}
#best fit plot####
plot(m4_bmass2, shade=T, pages=1)
```  

Fig.7. Partial effects plot of the relationships between different predictor variables and the proportion of breeding female PBs in k-rat exclosure plots estimated using a generalized additive model (GAM).  

**FOR DISCUSSION WITH MORGAN**: Same as in PB males in exclosure, model not necessarily with highest % deviance explained  

```{r message=FALSE, echo=FALSE, warning=FALSE}
want=seq(1, nrow(PB_female_ex), length.out = 200)
pdat2=with(PB_female_ex, data.frame(year=year[want], month=month[want], ndvi=ndvi[want], 
                                  precipitation=precipitation[want], bmass_DM=bmass_DM[want],
                                  bmass_PB=bmass_PB[want], bmass_PP=bmass_PP[want]))
p3=predict(m4_bmass, newdata=pdat2, type="terms", se.fit = TRUE)
pdat2=transform(pdat2, p3=p3$fit[,1], se3=p3$se.fit[,1]) #p2=fit, se2=std.error

df.res=df.residual(m4_bmass2)

crit.t=qt(0.025, df.res, lower.tail = F)
pdat2=transform(pdat2, upper=p3+(crit.t*se3), lower=p3-(crit.t*se3))

m4.d <- Deriv(m4_bmass2)
Term="month"
m4.dci <- confint(m4.d, term = Term)
m4.dsig <- signifD(pdat2$p3, d = m4.d[[Term]]$deriv,
                   +m4.dci[[Term]]$upper, m4.dci[[Term]]$lower)
plot.Deriv(m4.d, sizer=T, term=Term, xaxt="n")
axis(1, at=1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
abline(v=4.7, type="l", lty=2)

```  

Fig.8. Periods of significant change in the seasonal trends in the reproductive phenology of breeding female Bailey's pocket mouse (*Chaetodipus baileyi*) in k-rat exclosure plots in a project site near Portal, Arizona. Periods of significant increase (blue) or decrease (red) are indicated by the coloured sections of the trend. Dashed vertical line indicate peak breeding period.  

### PP MODELS  
```{r message=FALSE, include=FALSE}
PP_male_con=pp_plot%>%filter(treatment=="control", sex=="male")
PP_male_ex=pp_plot%>%filter(treatment=="exclosure", sex=="male")
PP_female_con=pp_plot%>%filter(treatment=="control", sex=="female")
PP_female_ex=pp_plot%>%filter(treatment=="exclosure", sex=="female")
```
#### PP MALES IN CONTROL   

```{r message=FALSE, echo=FALSE}
#no explicit biotic effects
m5=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)
             , data=PP_male_con, method = 'REML', weights = abundance, family = binomial)

summary(m5) #R2=54.5%

#add effect of DM+PB biomass
m5_bmass=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+
                     s(bmass_PB),
                   data=PP_male_con, method = 'REML', weights = abundance, family = binomial)

summary(m5_bmass) #R2=54.8%

#add effect of DM+PB biomass AND intraspecific competition (PP competition)
m5_bmass2=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+
                      s(bmass_PB)+s(bmass_PP), data=PP_male_con, method = 'REML', weights = abundance, family = binomial)

summary(m5_bmass2) #R2=55.7%
```  

```{r message=FALSE, echo=FALSE}
#best fit plot####
plot(m5_bmass2, shade=T, pages=1)

```  

Fig.9. Partial effects plot of the relationships between different predictor variables and the proportion of breeding male PPs in control plots estimated using a generalized additive model (GAM).  

```{r message=FALSE, echo=FALSE, warning=FALSE}
want=seq(1, nrow(PP_male_con), length.out = 200)
pdat2=with(PP_male_con, data.frame(year=year[want], month=month[want], ndvi=ndvi[want], 
                                    precipitation=precipitation[want], bmass_DM=bmass_DM[want],
                                    bmass_PB=bmass_PB[want],bmass_PP= bmass_PP[want]))
p3=predict(m5_bmass2, newdata=pdat2, type="terms", se.fit = TRUE)
pdat2=transform(pdat2, p3=p3$fit[,1], se3=p3$se.fit[,1]) #p2=fit, se2=std.error

df.res=df.residual(m5_bmass2)

crit.t=qt(0.025, df.res, lower.tail = F)
pdat2=transform(pdat2, upper=p3+(crit.t*se3), lower=p3-(crit.t*se3))

m5.d <- Deriv(m5_bmass2)
Term="month"
m5.dci <- confint(m5.d, term = Term)
m5.dsig <- signifD(pdat2$p3, d = m5.d[[Term]]$deriv,
                   +m5.dci[[Term]]$upper, m5.dci[[Term]]$lower)
plot.Deriv(m5.d, sizer=T, term=Term, xaxt="n")
axis(1, at=1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
abline(v=4.5, type="l", lty=2)

```  

Fig.10. Periods of significant change in the seasonal trends in the reproductive phenology of breeding male desert pocket mouse (*C. penicillatus*) in control plots in a project site near Portal, Arizona. Periods of significant increase (blue) or decrease (red) are indicated by the coloured sections of the trend. Dashed vertical line indicate peak breeding period.  

#### PP MALES IN EXCLOSURE  

```{r message=FALSE, echo=FALSE}  
#no explicit biotic effects
m6=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)
             , data=PP_male_ex, method = 'REML', weights = abundance, family = binomial)

summary(m6) #R2=59.6%

#add effect of DM+PB biomass
m6_bmass=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_PP),
                   data=PP_male_ex, method = 'REML', weights = abundance, family = binomial)

summary(m6_bmass) #R2=59.4%

#add effect of DM+PB biomass AND intraspecific competition (PB competition)
m6_bmass2=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+
                      s(bmass_PB)+s(bmass_PP), data=PP_male_ex, method = 'REML', weights = abundance, family = binomial)

summary(m6_bmass2) #R2=61%
```    
**FOR DISCUSSION WITH MORGAN**: As in male and female PPs in exclosure, not really best fit model chosen

```{r message=FALSE, echo=FALSE}

#best fit plot####
plot(m6_bmass2, shade=T, pages=1)
```  

Fig.11. Partial effects plot of the relationships between different predictor variables and the proportion of breeding male PPs in k-rat exclosure plots estimated using a generalized additive model (GAM).  


```{r message=FALSE, echo=FALSE, warning=FALSE}
want=seq(1, nrow(PB_male_ex), length.out = 200)
pdat2=with(PB_male_ex, data.frame(year=year[want], month=month[want], ndvi=ndvi[want], 
                                    precipitation=precipitation[want], bmass_DM=bmass_DM[want],
                                    bmass_PB=bmass_PB[want], bmass_PP=bmass_PP[want]))
p3=predict(m6_bmass, newdata=pdat2, type="terms", se.fit = TRUE)
pdat2=transform(pdat2, p3=p3$fit[,1], se3=p3$se.fit[,1]) #p2=fit, se2=std.error

df.res=df.residual(m6_bmass)

crit.t=qt(0.025, df.res, lower.tail = F)
pdat2=transform(pdat2, upper=p3+(crit.t*se3), lower=p3-(crit.t*se3))

m6.d <- Deriv(m6_bmass)
Term="month"
m6.dci <- confint(m6.d, term = Term)
m6.dsig <- signifD(pdat2$p3, d = m6.d[[Term]]$deriv,
                   +m6.dci[[Term]]$upper, m6.dci[[Term]]$lower)
plot.Deriv(m6.d, sizer=T, term=Term, xaxt="n")
axis(1, at=1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
abline(v=4.1, type="l", lty=2)
```  

Fig.12. Periods of significant change in the seasonal trends in the reproductive phenology of breeding male desert pocket mouse (*C. penicillatus*) in k-rat exclosure plots in a project site near Portal, Arizona. Periods of significant increase (blue) or decrease (red) are indicated by the coloured sections of the trend. Dashed vertical line indicate peak breeding period. 

#### PP FEMALES IN CONTROL  

```{r message=FALSE, echo=FALSE}
#no explicit biotic effects
m7=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)
             , data=PP_female_con, method = 'REML', weights = abundance, family = binomial)

summary(m7) #R2=51.9%

#add effect of DM+PB biomass
m7_bmass=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_PP),
                   data=PP_female_con, method = 'REML', weights = abundance, family = binomial)

summary(m7_bmass) #R2=52.8%

#add effect of DM+PB biomass AND intraspecific competition (PB competition)
m7_bmass2=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+
                      s(bmass_PB)+s(bmass_PP), data=PP_female_con, method = 'REML', weights = abundance, family = binomial)

summary(m7_bmass2) #R2=63.9% 
```  

```{r message=FALSE, echo=FALSE}
#best fit plot####
plot(m7_bmass2, shade=T, pages=1)
```  

Fig. 13. Partial effects plot of the relationships between different predictor variables and the proportion of breeding female PPs in control plots estimated using a generalized additive model (GAM).

```{r message=FALSE, echo=FALSE,warning=FALSE}
want=seq(1, nrow(PP_female_con), length.out = 200)
pdat2=with(PP_female_con, data.frame(year=year[want], month=month[want], ndvi=ndvi[want], 
                                   precipitation=precipitation[want], bmass_DM=bmass_DM[want],
                                   bmass_PB=bmass_PB[want],bmass_PP= bmass_PP[want]))
p3=predict(m7_bmass2, newdata=pdat2, type="terms", se.fit = TRUE)
pdat2=transform(pdat2, p3=p3$fit[,1], se3=p3$se.fit[,1]) #p2=fit, se2=std.error

df.res=df.residual(m7_bmass2)

crit.t=qt(0.025, df.res, lower.tail = F)
pdat2=transform(pdat2, upper=p3+(crit.t*se3), lower=p3-(crit.t*se3))

m7.d <- Deriv(m7_bmass2)
Term="month"
m7.dci <- confint(m7.d, term = Term)
m7.dsig <- signifD(pdat2$p3, d = m7.d[[Term]]$deriv,
                   +m7.dci[[Term]]$upper, m7.dci[[Term]]$lower)
plot.Deriv(m7.d, sizer=T, term=Term, xaxt="n")
axis(1, at=1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
abline(v=5.1, type="l", lty=2)
```  

Fig.14. Periods of significant change in the seasonal trends in the reproductive phenology of breeding female desert pocket mouse (*C. penicillatus*) in control plots in a project site near Portal, Arizona. Periods of significant increase (blue) or decrease (red) are indicated by the coloured sections of the trend. Dashed vertical line indicate peak breeding period. 

#### PP FEMALES IN EXCLOSURE  

```{r message=FALSE,echo=FALSE}
#no explicit biotic effects
m8=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)
             , data=PP_female_ex, method = 'REML', weights = abundance, family = binomial)

summary(m8) #R2=46.5%

#add effect of DM+PB biomass
m8_bmass=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_PP),
                   data=PP_female_ex, method = 'REML', weights = abundance, family = binomial)

summary(m8_bmass) #R2=52.4%

#add effect of DM+PB biomass AND intraspecific competition (PB competition)
m8_bmass2=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+
                      s(bmass_PB)+s(bmass_PP), data=PP_female_ex, method = 'REML', weights = abundance, family = binomial)

summary(m8_bmass2) #R2=61.7%
```

```{r message=FALSE, echo=FALSE}  

#best fit plot####
plot(m8_bmass2, shade=T, pages=1)
```  

Fig.15. Partial effects plot of the relationships between different predictor variables and the proportion of breeding female PPs in k-rat exclosure plots estimated using a generalized additive model (GAM).

```{r message=FALSE, echo=FALSE, warning=FALSE} 
want=seq(1, nrow(PP_female_ex), length.out = 200)
pdat2=with(PP_female_ex, data.frame(year=year[want], month=month[want], ndvi=ndvi[want], 
                                     precipitation=precipitation[want], bmass_DM=bmass_DM[want],
                                     bmass_PB=bmass_PB[want],bmass_PP= bmass_PP[want]))
p3=predict(m8_bmass2, newdata=pdat2, type="terms", se.fit = TRUE)
pdat2=transform(pdat2, p3=p3$fit[,1], se3=p3$se.fit[,1]) #p2=fit, se2=std.error

df.res=df.residual(m8_bmass)

crit.t=qt(0.025, df.res, lower.tail = F)
pdat2=transform(pdat2, upper=p3+(crit.t*se3), lower=p3-(crit.t*se3))

m8.d <- Deriv(m8_bmass2)
Term="month"
m8.dci <- confint(m8.d, term = Term)
m8.dsig <- signifD(pdat2$p3, d = m8.d[[Term]]$deriv,
                   +m8.dci[[Term]]$upper, m8.dci[[Term]]$lower)
plot.Deriv(m8.d, sizer=T, term=Term, xaxt="n")
axis(1, at=1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
abline(v=5.3, type="l", lty=2)

```  

Fig.16. Periods of significant change in the seasonal trends in the reproductive phenology of breeding female desert pocket mouse (*C. penicillatus*) in k-rat exclosure plots in a project site near Portal, Arizona. Periods of significant increase (blue) or decrease (red) are indicated by the coloured sections of the trend. Dashed vertical line indicate peak breeding period. 

### DM MODELS  
```{r message=FALSE, include=FALSE}
DM_male_con=dm_plot%>%filter(treatment=="control", sex=="male")
DM_male_ex=dm_plot%>%filter(treatment=="exclosure", sex=="male")
DM_female_con=dm_plot%>%filter(treatment=="control", sex=="female")
DM_female_ex=dm_plot%>%filter(treatment=="exclosure", sex=="female")
```  

#### DM MALES IN CONTROL 

```{r message=FALSE, echo=FALSE}

#no explicit biotic effects
m9=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)
             , data=DM_male_con, method = 'REML', weights = abundance, family = binomial)

summary(m9) #R2=18.4%

#add effect of DM+PB biomass
m9_bmass=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+
                     s(bmass_DM),
                   data=DM_male_con, method = 'REML', weights = abundance, family = binomial)

summary(m9_bmass) #R2=19.2%

#add effect of DM+PB biomass AND intraspecific competition (PP competition)
m9_bmass2=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+
                      s(bmass_PB)+s(bmass_PP), data=DM_male_con, method = 'REML', weights = abundance, family = binomial)

summary(m9_bmass2) #R2=22.1%
```  
**FOR DISCUSSION WITH MORGAN**: The DM models seem exhibit poor explanatory performance (<20% deviance explained for the best fit model). best fit shows DM and PP biomass influences P(breeding males), which is contrary to our hypothesis given DM's low density-dependence population regulation (also, with a lesser competitor). so I'm pretty skeptical about this result  

```{r message=FALSE, echo=FALSE}

#best fit plot####
plot(m9_bmass2, shade=T, pages=1)
```  

Fig.17. Partial effects plot of the relationships between different predictor variables and the proportion of breeding male DMs in control plots estimated using a generalized additive model (GAM).

```{r message=FALSE, echo=FALSE, warning=FALSE}
want=seq(1, nrow(DM_male_con), length.out = 200)
pdat2=with(DM_male_con, data.frame(year=year[want], month=month[want], ndvi=ndvi[want], 
                                   precipitation=precipitation[want], bmass_DM=bmass_DM[want],
                                   bmass_PB=bmass_PB[want],bmass_PP= bmass_PP[want]))
p3=predict(m9_bmass2, newdata=pdat2, type="terms", se.fit = TRUE)
pdat2=transform(pdat2, p3=p3$fit[,1], se3=p3$se.fit[,1]) #p2=fit, se2=std.error

df.res=df.residual(m9_bmass2)

crit.t=qt(0.025, df.res, lower.tail = F)
pdat2=transform(pdat2, upper=p3+(crit.t*se3), lower=p3-(crit.t*se3))

m9.d <- Deriv(m9_bmass2)
Term="month"
m9.dci <- confint(m9.d, term = Term)
m9.dsig <- signifD(pdat2$p3, d = m9.d[[Term]]$deriv,
                   +m9.dci[[Term]]$upper, m9.dci[[Term]]$lower)
plot.Deriv(m9.d, sizer=T, term=Term, xaxt="n")
axis(1, at=1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
abline(v=3.8, type="l", lty=2)

```  

Fig.18. Periods of significant change in the seasonal trends in the reproductive phenology of breeding male kangaroo rat (*Dipodomys merriami*) in control plots in a project site near Portal, Arizona. Periods of significant increase (blue) or decrease (red) are indicated by the coloured sections of the trend. Dashed vertical line indicate peak breeding period. 

#### DM FEMALES IN CONTROL  

```{r message=FALSE, echo=FALSE}
#no explicit biotic effects
m10=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)
             , data=DM_female_con, method = 'REML', weights = abundance, family = binomial)

summary(m10) #R2=39.1%

#add effect of DM+PB biomass
m10_bmass=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM),
                   data=DM_female_con, method = 'REML', weights = abundance, family = binomial)

summary(m10_bmass) #R2=39.1%

#add effect of DM+PB biomass AND intraspecific competition (PB competition)
m10_bmass2=mgcv::gam(proportion~s(month, bs="cc", k=12)+s(year)+s(ndvi)+s(precipitation)+s(bmass_DM)+
                      s(bmass_PB)+s(bmass_PP), data=DM_female_con, method = 'REML', weights = abundance, family = binomial)

summary(m10_bmass2) #R2=47.8%
```
**FOR DISCUSSION WITH MORGAN**: Not as terrible explanatory performance as DM males but surprising results: significant effect of PB and PP biomass on breeding females DMs  

```{r message=FALSE, echo=FALSE}
#best fit plot####
plot(m10_bmass2, shade=T, pages=1)
```  

Fig.19. Partial effects plot of the relationships between different predictor variables and the proportion of breeding female DMs in control plots estimated using a generalized additive model (GAM).

```{r message=FALSE, echo=FALSE, include=FALSE, warning=FALSE}
want=seq(1, nrow(DM_female_con), length.out = 200)
pdat2=with(DM_female_con, data.frame(year=year[want], month=month[want], ndvi=ndvi[want], 
                                   precipitation=precipitation[want], bmass_DM=bmass_DM[want],
                                   bmass_PB=bmass_PB[want],bmass_PP= bmass_PP[want]))
p3=predict(m10_bmass2, newdata=pdat2, type="terms", se.fit = TRUE)
pdat2=transform(pdat2, p3=p3$fit[,1], se3=p3$se.fit[,1]) #p2=fit, se2=std.error

df.res=df.residual(m10_bmass2)

crit.t=qt(0.025, df.res, lower.tail = F)
pdat2=transform(pdat2, upper=p3+(crit.t*se3), lower=p3-(crit.t*se3))

m10.d <- Deriv(m10_bmass2)
Term="month"
m10.dci <- confint(m10.d, term = Term)
m10.dsig <- signifD(pdat2$p3, d = m10.d[[Term]]$deriv,
                   +m10.dci[[Term]]$upper, m10.dci[[Term]]$lower)
plot.Deriv(m10.d, sizer=T, term=Term, xaxt="n")
axis(1, at=1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sept", "Oct", "Nov", "Dec"))
abline(v=4.7, type="l", lty=2)
```  

Fig.20. Periods of significant change in the seasonal trends in the reproductive phenology of breeding female kangaroo rat (*Dipodomys merriami*) in control plots in a project site near Portal, Arizona. Periods of significant increase (blue) or decrease (red) are indicated by the coloured sections of the trend. Dashed vertical line indicate peak breeding period. 

### **NEXT STEPS (SUMMER TO DO LIST)**  
* do manual checking of breeding data  
* figure out a way to overlay the phenology plot with the first derivatives plot to make it more intuitive 
* figure out an efficient way to present results to reduce number of figures (20 figures is preposterous, Pat!)  
* determine how important standardizing variables are in building GAMs and interpreting them  
* read Gavin's papers to see if there are more info. we can extract from GAM outputs beyond linear/non-linear relationships  
* *challenge*: implement the model in a Bayesian framework and see if the model output differs  (use brms?) 

