---
title: "Phenological traits of Rodents in Portal, AZ (male)"
author: "Pat Dumandan"
date: "11/23/2020"
output: html_document
---

### Counts of individuals in S,M,R in set-ups
#### Exclosure set-up
```{r, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
Portal_rodent=read.csv("./PortalData/Rodents/Portal_rodent.csv")
Portal_plots=read.csv("./PortalData/SiteandMethods/Portal_plots.csv")
portal1=inner_join(Portal_rodent, Portal_plots, by= c("plot", "month", "year"))
```

```{r, message=FALSE, warning=FALSE, }
target_sp=c("PB", "PP", "PE", "PF", "PM", "RM")
target_repro=c("S", "M", "R")
target_month=c(1:12)
total_ex=portal1%>%
  filter(species %in% target_sp, testes %in% target_repro,
         treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)
#write.csv(total_ex, "male_exclosure_SMR.csv")
```
*if only S in exclosure=1838 (vs 3008 if including M and R)*

##### Control set-up
```{r}
total_con=portal1%>%
  filter(species %in% target_sp, testes %in% target_repro,
         treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)
#write.csv(total_con, "male_control_SMR.csv")
```
*if only S in control=1333 (vs 2320 if including M and R)*

### Proportion of counts per month per species (scrotal only)
```{r, message=FALSE, warning=FALSE}
#control with scrotal only

#sp1: PB
count_per_month_PBcon= c(4,17,34,19,12,9,7,4,2,5,2,4)
count_per_month_PBex= c(6,14,47,46,34,22,27,10,11,14,3,7)
PBprop=vector(mode="integer", length=length(count_per_month_PBcon))
PBpropex=vector(mode="integer", length=length(count_per_month_PBex))

for (i in 1:length(count_per_month_PBex)) {
  proportions=count_per_month_PBex[i]/sum(count_per_month_PBex)
  PBpropex[i]=proportions
}
PBpropex

for (i in 1:length(count_per_month_PBcon)) {
  proportions=count_per_month_PBcon[i]/sum(count_per_month_PBcon)
  PBprop[i]=proportions
}

PBprop

```

```{r, message=FALSE, warning=FALSE}
# control with S,M,R 
PBcon=portal1%>%
  filter(species=="PB", testes %in% target_repro,
         treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)

PBex=portal1%>%
  filter(species=="PB", testes %in% target_repro,
         treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)

count_per_month_PBcon_smr= c(27,58,95,49,49,35,27,20,21,17,26,21)
count_per_month_PBex_smr= c(43,66,136,95,95,79,73,48,57,47,40,30)
PBprop_smr=vector(mode="integer", length=length(count_per_month_PBcon_smr))
PBpropex_smr=vector(mode="integer", length=length(count_per_month_PBex_smr))

for (i in 1:length(count_per_month_PBcon_smr)) {
  proportions=count_per_month_PBcon_smr[i]/sum(count_per_month_PBcon_smr)
  PBprop_smr[i]=proportions
}
PBprop_smr

for (i in 1:length(count_per_month_PBex_smr)) {
  proportions=count_per_month_PBex_smr[i]/sum(count_per_month_PBex_smr)
  PBpropex_smr[i]=proportions
}
PBpropex_smr
```
*PBcontrol=around 69% of the total "events" occurred in the months of FEB-MAY (~ >10%/month proportions)* 

*PBexclosure=around 72% of the total "events" occurred in the months of MAR-MAY (~ >10%/month proportions)*   

*PB exclosure with all events(SMR), bulk of these events (51%) occurred in MAR-JUN (~ >10%/month proportions)*    

*PB control with all events(SMR), bulk of these events (56%) occurred in FEB-MAY (~ >10%/month proportions)*   

*Qualitatively, it seems that for PBs, the reproductive events (across all years) tend to occur earlier in the control set-ups (Feb-May)*  

### Proportion of individuals in reproductive state per month per year
```{r, message=FALSE, warning=FALSE}
#EXCLOSURES#####
#total number of male PBs
PBall=portal1%>%
  filter(species=="PB",
         treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)
length(PBall$month) #5711

#number of male PBs in reproductive state
PBex=portal1%>%
  filter(species=="PB", testes %in% target_repro,
         treatment== "exclosure", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)
length(PBex$month) #809

#proportion of male PBs in reproductive state per month across all years
#exclosures

#HAVE TO FIGURE OUT A WAY TO AUTOMATE THIS USING A FOR LOOP~!
mon=PBall%>%
    filter(month==12)

count_per_month_PBall_ex= c(328,336,508,321,476,585,634,453,573,494,533,470)
count_per_month_PBex_smr= c(43,66,136,95,95,79,73,48,57,47,40,30)

PBex_prop=vector(mode="integer", length=length(count_per_month_PBall_ex))

for (i in 1:length(count_per_month_PBall_ex)) {
  proportions=count_per_month_PBex_smr[i]/sum(count_per_month_PBall_ex)
  PBex_prop[i]=proportions
}
sum(PBex_prop)


#CONTROL
#total number of male PBs
PBallcon=portal1%>%
  filter(species=="PB",
         treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)
length(PBallcon$month) #2808

#number of male PBs in reproductive state
PBcon=portal1%>%
  filter(species=="PB", testes %in% target_repro,
         treatment== "control", !(age=="J"))%>%
  select(month, day, year, treatment, species, testes)
length(PBcon$month) #445

#proportion of male PBs in reproductive state per month across all years
#control

#HAVE TO FIGURE OUT A WAY TO AUTOMATE THIS USING A FOR LOOP~!
mon=PBallcon%>%
    filter(month==12)

count_per_month_PBall_con= c(172,187,284,158,223,281,307,210,274,227,249,236)
count_per_month_PBcon_smr= c(27,58,95,49,49,35,27,20,21,17,26,21)

PBcon_prop=vector(mode="integer", length=length(count_per_month_PBall_con))

for (i in 1:length(count_per_month_PBall_con)) {
  proportions=count_per_month_PBcon_smr[i]/sum(count_per_month_PBall_con)
  PBcon_prop[i]=proportions
}
PBcon_prop
```
*qualitatively, Feb-Jul are the months when more male individuals are in reproductive state across all years in exclosures (>1%)*  
*(809/5711)X100= only 14% of the total males were in reproductive state across all years in exclosure*
*qualitatively, Feb-Jun are the months when more male individuals are in reproductive state across all years in controls (>1%)* 
*(445/2808)X100= only ~16% of the total males were in reproductive state across all years in control*

### temporal patterns of proportion of indivs. in reproductive state

```{r, message=FALSE, warning=FALSE}
#### calculate proportion per month per year for PB male

#1: extract count of events per month
#totals per month: all
PBt=portal1%>%
  filter(species=="PB",
         treatment== "control", !(age=="J"))%>%
  group_by(year,month)%>%
  summarise(count=n())%>%
  select(month,year, count)
PBt

#totals per month: S
PBs=portal1%>%
  filter(species=="PB",
         treatment== "control", !(age=="J"), testes=="S")%>%
  group_by(year,month, testes)%>%
  summarise(count=n())%>%
  select(month,year, count, testes)
PBs

#2: apply to for loop: still have to figure this out
#a:for each year, get proportions per month

#for (i in 1: mon){
 # for (i in 1:Nyr){
  #  proportions[i]= count_per_yr_mon[i]/count_per_yr_tot[i]
  #}
count_per_month_PBall_con= c(172,187,284,158,223,281,307,210,274,227,249,236)
count_per_month_PBcon_smr= c(27,58,95,49,49,35,27,20,21,17,26,21)

PBcon_prop=vector(mode="integer", length=length(count_per_month_PBall_con))

for (i in 1:length(count_per_month_PBall_con)) {
  proportions=count_per_month_PBcon_smr[i]/sum(count_per_month_PBall_con)
  PBcon_prop[i]=proportions
}
```

#### **to ask/discuss with Morgan:**   
* is it worth it to add M and R given these counts?  
* is it worthwhile to ask whether the chronology of the reproductive events of species follows some order based on size, diet preference, etc.?/relate it with other traits?  
* peak timing: across all years, in which months do we observe the highest proportion of reproductive events?  
* small sample sizes for each month per year has led me to use the proportion of species in reproductive state but have to figure out what to do in months where no observations for a given species were made in a given year

#### **Other stuff to ask Morgan:**  
* can I still go to the office over the holidays?  
* class:Biodiversity theory by Bob Holt?